---
title: "CHILD CBMC and age one PBMC deconvolution"
author: "SL"
date: "2020-10-27"
output: html_document
---

### Purpose 

Cell type deconvolution allows prediction of cell type proportions in cordblood and PMBCs. Cell type proprotions are corrected for in later analysis. 

*NB*: For deconvolution using estimateCellCounts2 or estimateCellCounts must use RGset (estimateCellCounts2 allows mset but functionality is reduced). CHILD rgset_no_badsamp_probe_lowbead has bad samples removed. Bad probes, cross reactive probes, xy probes, and SNP probes have not yet been removed. 

### Load libraries and data 

Load required libraries
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse) # for dplyr
library(minfi) # for dnam data
library(ExperimentHub) # for flow sorted sets
library(here) # for dir calling
library(FlowSorted.CordBloodCombined.450k)
library(FlowSorted.Blood.EPIC)
library(FlowSorted.CordBlood.450k)
```


Load rgset_no_badsamp_probe_lowbead for deconvolution comparison.  
```{r load_rgset}
load(here("output_data", 
          "preprocessed_data", 
          "2023-06-28_child_rgset_nobadsamp.Rdata"))
```


Separate cordblood (CMBCs) and year 1 PMBCs.
```{r separate_tissues}
# correct sample mix up in rgset
# sample mix up was determined in preprocessing and analysis after 
# the rgset was outputted
# tissue mix up for sample 20113 -> relabel
rgset_nobadsamp$SampleID[5] <- "20113_P"
rgset_nobadsamp$Tissue[5] <- "P"
rgset_nobadsamp$SampleID[6] <- "20113_C"
rgset_nobadsamp$Tissue[6] <- "C"

# cord blood 
cb_rgset <- rgset_nobadsamp[,rgset_nobadsamp$Tissue=="C"]
dim(cb_rgset) # 153

# year 1 pmbcs
y1_rgset <- rgset_nobadsamp[,rgset_nobadsamp$Tissue=="P"]
dim(y1_rgset) # 145
```

Load improved estimateCellCounts2 function. This also includes a line within the internal function pickCompProbes to output the probe list uses in cell type deconvolution.
Load improved estimated cell counts 2 function
```{r load_estimatecellcounts2}
source(here("scripts",
            "2022-09-27_estimateCellCounts2v2.R"))
```

Deconvolution of CBMCs (cord blood) and year 1 PBMCs using cordblood reference sets using estimateCellCounts2 and FlowSorted.CordBloodCombined.450k 
```{r child_fscbc_ecc2}
hub <- ExperimentHub()
myfiles <- query(hub, "FlowSorted.CordBloodCombined.450k")
FlowSorted.CordBloodCombined.450k <- myfiles[[1]]
FlowSorted.CordBloodCombined.450k

# CBMCs with cord blood reference set 
# dont include granulocytes since they are not expected in CBMCs
child_fscbc_ecc2_cb <- 
  estimateCellCounts2v2(rgSet = cb_rgset, 
                        compositeCellType = "CordBlood",
                        processMethod = "preprocessNoob",
                        probeSelect = "any",
                        cellTypes = c("CD4T","NK","nRBC","Bcell","Mono","CD8T", "Gran"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.CordBloodCombined.450k",
                        returnAll = FALSE) 

child_fscbc_ecc2_cb$probelist_all <- probelist_all
child_fscbc_ecc2_cb$probelist_deconvo_coeffEsts <- compData$coefEsts

# year 1 PBMCs with cordblood reference set
# dont include granulocytes or nRBCs since neither are expected in year 1 PBMCs
child_fscbc_ecc2_y1 <- 
  estimateCellCounts2v2(rgSet = y1_rgset, 
                        compositeCellType = "CordBlood",
                        processMethod = "preprocessNoob",
                        cellTypes = c("CD4T","NK", "Bcell","Mono","CD8T", "Gran"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.CordBloodCombined.450k",
                        returnAll = FALSE) 

child_fscbc_ecc2_y1$probelist_all <- probelist_all
child_fscbc_ecc2_y1$probelist_deconvo_coeffEsts <- compData$coefEsts
```

Save deconvoluted data.    
```{r deconvo_data}
save(child_fscbc_ecc2_cb, 
     child_fscbc_ecc2_y1, 
     file=here("output_data",
               "deconvolution", 
               "2023-08-14_cbmc_pbmc_deconvolution.Rdata"))

sesh <- sessionInfo()

save(sesh, 
     file=here("output_data",
               "deconvolution", 
               "2023-08-14_cbmc_pbmc_deconvolution_session_info.Rdata"))

# write as csvs for cell estimates data
child_fscbc_ecc2_cb_counts <- as.data.frame(child_fscbc_ecc2_cb$prop)
child_fscbc_ecc2_y1_counts <- as.data.frame(child_fscbc_ecc2_y1$prop)

write.csv(child_fscbc_ecc2_cb_counts, 
          here("tables",
               "deconvolution",
               "2023-08-14_cbmc_estimates.csv"))

write.csv(child_fscbc_ecc2_y1_counts, 
          here("tables",
               "deconvolution",
               "2023-08-14_pbmc_estimates.csv"))
```