---
title: "CHILD cord blood cell cell type PC generation"
author: "SL"
date: "2022-09-27"
output: html_document
---

*Purpose*: This script generates cell type PCs from CBMCs. CBMC PCs can be used in linear models to correct for cell type.

## Cell type correction

Correct cordblood for cell type using linear regression. 

Based on "Adjusting for cell type composition in DNA methylation data using a regression-based approach" by Jones et al (2017).  

The data is composite data and cell types are not independent of one another:  

Two ways of dealing with this:  

1. Do a PCA  
* variables are not independent -PCA will give all the same data but in independent vectors  
* But cant do PCA on composite data - this is composite data  
* Do a centred log ratio before PCA to get around this  

2. Leave out one cell type  
* More simple way  
* Usually leave cell type that is largest proportion  
* For whole blood/cord blood - drop granulocytes  
* For PMBCs and CBMCs - drop CD4T cells  


## Principle Component (PC) generation

Load required libraries.
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(minfi)
library(here)
library(compositions)
library(MASS)
```

Load CHILD145 deconvolution data. This deconvolution does not include granulocytes as granulocytes are not expected.
```{r load_CHILD_data, message=FALSE, warning=FALSE}
load(here("output_data", 
          "deconvolution", 
          "2023-06-28_cbmc_pbmc_deconvolution.Rdata"))

#rm year one data
rm(child_fscbc_ecc2_y1)
```


Remove duplicates from cord blood deconvolution data.  
```{r remove_dups_from_deconvo}
#convert deconvolution data to dataframe
child_fscbc_ecc2_cb_df <- as.data.frame(child_fscbc_ecc2_cb$prop)

# make list of samples to remove from beta and pData
dups_rm <-c("9298768102_R06C01", "9341679111_R05C02", "9298768023_R03C02",
            "9341679076_R06C02", "9297962089_R06C02", "9298768023_R02C02",
            "9341679097_R02C02", "9341679114_R01C02", "9298768023_R06C01")

# remove replicates from cell type estimates                      
child_fscbc_ecc2_cb_df_nodups <-
  child_fscbc_ecc2_cb_df[!rownames(child_fscbc_ecc2_cb_df) %in% dups_rm, ]

# check that 9 samples were removed
dim(child_fscbc_ecc2_cb_df_nodups) # 144 6
dim(child_fscbc_ecc2_cb_df) # 153 6
```

Use ILR and robust PCA as there are outliers in nRBCs
```{r ilr_robustPCA}
cb_ilr <- compositions::ilr(child_fscbc_ecc2_cb_df_nodups[,1:7])

cv_cb <- MASS::cov.rob(cb_ilr %>% as.data.frame() %>% as.matrix(), cor = FALSE)
cb_ilr_pca <- suppressWarnings(princomp(cb_ilr %>% as.data.frame() %>% as.matrix(), 
                                                 covmat = cv_cb, 
                                                 cor = FALSE))
```

Examine variance captured by PCs.
```{r pc_variance}
var_explained = cb_ilr_pca$sdev^2 /sum(cb_ilr_pca$sdev^2)
ggplot(var_explained %>% as.data.frame(), aes(y=., x=(c(1:6)))) + geom_col()
cumsum(var_explained) 
# Comp.1    Comp.2    Comp.3    Comp.4    Comp.5    Comp.6 
# 0.4726463 0.6508847 0.7946900 0.9218764 0.9755405 1.0000000
```

Save CBMC PCs. 
```{r save_adjusted_betas}
rownames(cb_ilr_pca$scores) <- child_fscbc_ecc2_cb_df_nodups %>% rownames()

save(cb_ilr_pca, 
     file=here("output_data", 
               "deconvolution", 
               "2023-08-14_CBMC_PCs.RData"))

sesh <- sessionInfo()

save(sesh, 
     file=here("output_data", 
               "deconvolution", 
               "2023-08-14_CBMC_PCs_session_info.RData"))

# write our cbmc PCs for 
write.csv(cb_ilr_pca$scores,
          here("tables",
               "deconvolution",
               "2023-08-14_cbmc_PCs.csv"))
```