---
title: "CHILD cord blood linear regression using limma with covariate missing values filled in with means"
author: "Samantha LEE"
date: "2021-09-24"
output: html_document
---

*Purpose*: The purpose of this script is to examine DNAm changes in cord blood associated with prenatal air pollution exposure (proxied by no2).


Set global chunk options. 
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load required libraries.
```{r libraries}
library(tidyverse) # be tidy
library(limma) # for DNAm differences
library(here) # file calling
library(pbapply) # for running lm for each cpg
library(ggeffects) # for estimating marginal effects
library(broom) # tidy function for summarizing lm
library(sva) # for calculating SVs
library(ggcorrplot) # correlation matrix
library(bacon) # pvalue correction
library(ggsignif) # for significance comparison bars
```


Load betas
```{r betas}
load(file=here("output_data", "preprocessed_data",
               "2021-10_07_CHILD_preprocessed_betas_pdata_annotation.Rdata"))
```


Subset out cord blood betas
```{r nocombat_betas}
# check ranges to confirm DNAm data is in beta value format
dim(betas)
max(betas)
min(betas)

# subset pData into CBMCs (cord blood)
# these children have data at both time points
cb_pdata <- subset(pdata_nodups, Tissue == "C")
# check size of cord blood pdata
dim(cb_pdata)

# subset cordb lood DNAm samples for those included in cord blood pdata
cb_betas <- betas[ ,colnames(betas) %in% rownames(cb_pdata)]

# check size for correct sample number
dim(cb_betas) #144

# check that CBMC betas and pdata are in same order
identical(colnames(cb_betas), rownames(cb_pdata)) # TRUE

# relabel CBMC beta columns with sample label (5 digit identifier)
colnames(cb_betas) <- cb_pdata$Sample_Label
```


Load covariate data. Missing values in covariates of interest have been filled with the mean of that variable. 
```{r covariates}
load(file=here("output_data", "covariates", "2021-09-24_CHILD_covariates_na_with_means.Rdata"))
```


Subset covariates for individuals with CBMC DNAm.
```{r sub_covariates}
# subset covariates based on samples with DNAm data
covariates_sub <- covariates_nameans[as.character(covariates_nameans$sampleid) 
                                     %in% colnames(cb_betas), ]

# check size
dim(covariates_sub) # 144 113  

# check if subset of covariates and betas are in the same order 
identical(as.character(covariates_sub$sampleid), colnames(cb_betas)) # TRUE
```


Load CBMC PCs.
```{r load_PC_correctedbetas}
# load the cell type PCs
load(here("output_data", "deconvolution", "2021-09-24_CBMC_PCs.RData"))

# check size
dim(cb_pcs) #144 6

#check if rownames of cell PCs are same as pdata
identical(rownames(cb_pcs), rownames(cb_pdata)) # TRUE

# rename rows using sample label (5 digit number)
rownames(cb_pcs) <- cb_pdata$Sample_Label

# rename columns for clarity
colnames(cb_pcs) <- paste("cbmc.", colnames(cb_pcs), sep="")
```


Load genotyping PCs and subset out REEGLE PCs. 
Genotyping data contains PCs for all CHILD participants that were genotyped.
```{r genotyping_PCS}
# read in genotyping data for CHILD cohort
genotyping <- read.csv(here("input_data", "genotyping", "CHILD_subjects_Qced_first10PCS_Duan.csv"))

# make rownames sample names (5 digit number)
rownames(genotyping) <- genotyping$FID
# removing sample naming columns - no longer needed now they are rownames 
genotyping <- subset(genotyping, select=-c(FID, IID))

# susbet genotyping data for REEGLE subset 
genosub <- genotyping[rownames(genotyping) %in% cb_pdata$Sample_Label,]

# check size
dim(genosub) # 131 of 144 children have genotyping

# rename columns of genotyping data for clarity
colnames(genosub) <- paste("genotyping.", colnames(genosub), sep="")

#put rows in numerical order according to sample id
genosub_order <- genosub[order(rownames(genosub)), ]
```


Examine genetic ancestry groups of participants based on genotyping PCs. Want to confirm that three PCs are required for genetic ancestry.
```{r genotyping_pc_groups, eval=FALSE}
# load genotyping info from Amirtha/Qingling 
# this info is for all CHILD participants that were genotyped
pca_info <- readxl::read_excel(here("input_data", "genotyping", "PCA_ethnicity_child_comparison_duan_lab07102020.xlsx"))

# sub for individuals in this study (REEGLE subset)
pca_sub <- pca_info[as.character(pca_info$FID) %in% rownames(genosub),]

# plot and colour according to race specified on questionnaires
ggplot(pca_sub, aes(x=PC1, y=PC2, 
                    col=as.factor(race_mom), 
                    shape=as.factor(`PCA ethnicity`))) +
  geom_point(alpha=0.7, size=2) +
  scale_color_manual(labels = c("First Nations", "Mixed","South East Asian",
                                "East Asian","South Asian","Middle Eastern",
                                "Hispanic","Caucasian White"),
                     values=c("grey55","#4E2A7A" , "#B20D30","#C17817",
                              "#3F784C", "black", "#6DB1BF","#254A90"),
                     name ="Maternal race")  +
  scale_shape_manual(values=c(21,22,23,24),
                     labels=c("Admixed","Central European" , 
                              "East Asian","South Asian"),
                     name ="PCA Genetic Ancestry") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(size=6),
        axis.text.y = element_text(size=6),
        axis.title.x = element_text(size=8),
        axis.title.y = element_text(size=8),
        legend.text=element_text(size=4),
        legend.title=element_text(size=5),
        legend.key.size = unit(0.5, 'cm'), #change legend key size
        legend.key.height = unit(0.3, 'cm'), #change legend key height
        legend.key.width = unit(0.5, 'cm')) 

#  ggsave( filename="2021-01-21_CHILD_genopcs_race_ancestry.tiff",
#          device = "tiff",
#          path =  here("figures", "genotyping"),
#          width = 3.5,
#          height = 2.5,
#          units = "in",
#          limitsize = TRUE)
```


Subset betas, pdata, covariates, and cell type PCS for participants with genotyping data (the limiting factor).
```{r subset}
# subset betas
cb_betas_geno <- cb_betas[, rownames(genosub_order)]
# check size of subbed betas
dim(cb_betas_geno) # 424644    131

# subset pdata
cb_pdata_geno <- cb_pdata[cb_pdata$Sample_Label %in% rownames(genosub_order), ]
# check size of subbed pdata
dim(cb_pdata_geno) # 131 17

# subset covariates
covariates_sub_geno <- covariates_sub[covariates_sub$sampleid %in% 
                                        rownames(genosub_order), ]
# check size of subbed covariates 
dim(covariates_sub_geno) # 131 114

# subset cell type PCs
cb_pcs_geno <- cb_pcs[rownames(genosub_order), ]
# check size of subbed cell type PCs
dim(cb_pcs_geno) # 131 10
```


Check that betas, pdata, covariates, cell type PCs are in same order as genotyping data for REEGLE participants.
```{r cb_batchcorr_order}
# pdata and genosub_order
identical(as.character(cb_pdata_geno$Sample_Label),rownames(genosub_order)) # TRUE

# betas and genosub_order
identical(colnames(cb_betas_geno), rownames(genosub_order)) # TRUE

# covariates and genosub_order
identical(as.character(covariates_sub_geno$sampleid), rownames(genosub_order)) # TRUE

# cell type PCs and genosub_order
identical(rownames(genosub_order), rownames(cb_pcs_geno)) # TRUE
```


Combined covariates, genotyping, cell type PCs, and pData together.
```{r combine_cell_pcs_covariates}

###################################
# clean up pdata before combining #
###################################

# create new column for batch row variable in pdata
cb_pdata_geno$row <- as.numeric(as.factor(substr(cb_pdata_geno$Sentrix_Position,
                                                 1, 3)))
# examine
head(cb_pdata_geno$row)

# rename pdata sentrix Id column to chip
colnames(cb_pdata_geno)[8] <- "chip"

# rename pdata sentrix Id column to chip
colnames(cb_pdata_geno)[11] <- "run"

################################
# combine data frames together #
################################

# for pdata, only include batch variables
covariates_geno_ctpc_tech <- cbind(covariates_sub_geno, 
                                   cb_pcs_geno,
                                   genosub_order,
                                   cb_pdata_geno[,c("run", "row", "chip")])

# check classes to make sure they are what we expect before running linear model
lapply(covariates_geno_ctpc_tech, class)

# reclass run 
covariates_geno_ctpc_tech$run <- factor(covariates_geno_ctpc_tech$run, 
                                        ordered=F)
# reclass chip
covariates_geno_ctpc_tech$chip <- factor(covariates_geno_ctpc_tech$chip,
                                         ordered=F)
# reclass sex
covariates_geno_ctpc_tech$sex <- factor(covariates_geno_ctpc_tech$sex, 
                                        ordered=F)
```


Remove rows where air pollution is 0.
```{r no_zeroes_in_no2}
covariates_geno_ctpc_tech_not0 <- covariates_geno_ctpc_tech[!covariates_geno_ctpc_tech$no2_preg_entire==0, ]

dim(covariates_geno_ctpc_tech) # 131 132
dim(covariates_geno_ctpc_tech_not0) # 125 132
```


Subset complete cases in covariate data and betas in preparation for DNAm analysis.
```{r complete_cases}
# subset complete cases
covariates_geno_ctpc_tech_cc <- covariates_geno_ctpc_tech_not0 %>%
  subset(complete.cases(no2_preg_entire, sex, gestational_days,
                        prenatal_maternal_smoking, maternal_education_length,
                        pss_18wk, cbmc.PC1, cbmc.PC2, cbmc.PC3, cbmc.PC4,
                        genotyping.PC1, genotyping.PC2, genotyping.PC3))

# check number of complete cases
dim(covariates_geno_ctpc_tech_cc) # 95 132

# subset beta values for complete cases
cb_betas_geno_cc <- cb_betas_geno[,colnames(cb_betas_geno) %in% 
                                    covariates_geno_ctpc_tech_cc$sampleid]

# check size of complete case betas
dim(cb_betas_geno_cc) # 424644 95

# remove probes from betas that were used to predict cell type
cb_betas_geno_cc_clean <- 
  cb_betas_geno_cc[!rownames(cb_betas_geno_cc) %in%
                     rownames(child_fscbc_ecc2_cb_probelist_deconvo), ]

# check size
dim(cb_betas_geno_cc_clean) # 424079 95; 565 probes removed
```


Create model matrix for DNAm analysis. This model matrix does not contain batch variables as we are going to let surrogate variables capture their variance within the DNAm data. 
```{r model_matrix}
# create model matrix
mod <- model.matrix(~ no2_preg_entire + sex + gestational_days + pss_18wk + 
                      prenatal_maternal_smoking + maternal_education_length +
                      cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  +
                      genotyping.PC1 + genotyping.PC2 + genotyping.PC3,
                    data=covariates_geno_ctpc_tech_cc)

# check size
dim(mod) # 95 15
```


Surrogate variable analysis using the sva packages. This will allow us to capture variance due to batch effects, as well as variance due to unkown biological variables. 
```{r sva}

###############
# null matrix #
###############

# create null matrix
# this matrix does not include any of our covariates 
# this means that our covariates are "protected" and SVs will not capture them
mod0 <- model.matrix(~ sex + gestational_days + pss_18wk + 
                       prenatal_maternal_smoking + maternal_education_length +
                       cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                       genotyping.PC1 + genotyping.PC2 + genotyping.PC3,  
                     data=covariates_geno_ctpc_tech_cc)

################
# sva analysis #
################

# to get SVs in a way that allows calculation of their variance
# we need to pull the code out of sva and run it line-by-line
# we are using the "Be" method for calculating SVs

######################################
# irwsva required internal functions #
######################################

# edge.lfdr function is internal to irwsva function below
# mono function is internal to edge.lfdr
# they have to be initialized before we can run the rest of the code

set.seed(1234)

mono <- function(lfdr){
  .Call("monotone", as.numeric(lfdr), PACKAGE="sva")
}

edge.lfdr <- function(p, trunc=TRUE, monotone=TRUE, transf=c("probit", "logit"), adj=1.5, eps=10^-8, lambda=0.8, ...) {
  pi0 <- mean(p >= lambda)/(1 - lambda)
  pi0 <- min(pi0, 1)
  n <- length(p)
  transf <- match.arg(transf)
  if(transf=="probit") {
    p <- pmax(p, eps)
    p <- pmin(p, 1-eps)
    x <- qnorm(p)
    myd <- density(x, adjust=adj)
    mys <- smooth.spline(x=myd$x, y=myd$y)
    y <- predict(mys, x)$y
    lfdr <- pi0*dnorm(x)/y
  }
  if(transf=="logit") {
    x <- log((p+eps)/(1-p+eps))
    myd <- density(x, adjust=adj)
    mys <- smooth.spline(x=myd$x, y=myd$y)
    y <- predict(mys, x)$y
    dx <- exp(x) / (1+exp(x))^2
    lfdr <- pi0 * dx/y
  }
  if(trunc) {
    lfdr[lfdr > 1] <- 1
  }
  if(monotone) {	
    lfdr <- lfdr[order(p)]
    lfdr <- mono(lfdr)
    lfdr <- lfdr[rank(p)]
  }
  return(lfdr)
}



##################################
# numsv function internal to sva #
##################################

#determine the number of required SVs
#this is numsv function code pulled out

#this is what is fed into this function from the sva function
dat = cb_betas_geno_cc_clean
mod = mod
#method="be" # this is the default and what is called by numsv
#vfilter=NULL # we dont apply a filter and its not used for be method
B=20
set.seed(1234)

# initialize some stuff
# get the residual matrix (the remaining variance)
n <- ncol(dat)
m <- nrow(dat)
H <- mod %*% solve(t(mod) %*% mod) %*% t(mod)
res <- dat - t(H %*% t(dat))

# run svd and get some eigenvalues
uu <- svd(res)
# get the num of "sig" eigenvalues
ndf <- min(m, n) - ceiling(sum(diag(H)))
# get percent variance explained by each eigenvalue
dstat <- uu$d[1:ndf]^2/sum(uu$d[1:ndf]^2)
# initialize null distribution matrix
dstat0 <- matrix(0, nrow = B, ncol = ndf)

# set seed here since we are calling sample
# create the null distribution
for (i in 1:B) {
  # this part of the code removes all relationships in the 
  res0 <- t(apply(res, 1, sample, replace = FALSE))
  res0 <- res0 - t(H %*% t(res0))
  # calculate svs on this
  uu0 <- svd(res0)
  # bind their explained variance in null matrix
  dstat0[i, ] <- uu0$d[1:ndf]^2/sum(uu0$d[1:ndf]^2)
  print(i) # print out the iteration we are on
}

# determine the number of svs required here based on pvalues
psv <- rep(1, n)
for (i in 1:ndf) {
  psv[i] <- mean(dstat0[, i] >= dstat[i])
}
for (i in 2:ndf) {
  psv[i] <- max(psv[(i - 1)], psv[i])
}

# this is the variable holding the number of SVs required 
# number is each to all the svs with a pval less than/equal to 0.1
nsv <- sum(psv <= 0.1)


####################################
# code from the main SVA function  #
####################################

n.sv = nsv # from code above (run internally when sva called)
B = 5 # number of iterations

# dat =  betas # already initialized
# mod = mod  # already initialized
# mod0 = mod0  # already initialized 
# controls = NULL # we dont include controls - only for supervised method
# method = "irw" # this isnt actually used but would be called by sva
# vfilter = NULL # we dont set this when running sva
# numSVmethod = "be" # this is the default called by sva (code above)


###################################
# irwsva function internal to sva #
###################################

# this code is really want does all the work in sva
# we use the irw method so this is the code for that method
# there are also supervised and two step methods (not used by us)

# irw function takes input from sva, so we have already defined below vars
# dat = dat 
# mod = mod
# mod0 = mod0
# n.sv = n.sv
# B = B

# intialize variables again
n <- ncol(dat)
m <- nrow(dat)
Id <- diag(n)
resid <- dat %*% (Id - mod %*% solve(t(mod) %*% mod) %*% t(mod))
uu <- eigen(t(resid) %*% resid)
vv <- uu$vectors
ndf <- n - dim(mod)[2]
pprob <- rep(1, m)
one <- rep(1, n)
Id <- diag(n)
df1 <- dim(mod)[2] + n.sv
df0 <- dim(mod0)[2] + n.sv
rm(resid)

cat(paste("Iteration (out of", B, "):"))
for (i in 1:B) {
  mod.b <- cbind(mod, uu$vectors[, 1:n.sv])
  mod0.b <- cbind(mod0, uu$vectors[, 1:n.sv])
  ptmp <- f.pvalue(dat, mod.b, mod0.b)
  pprob.b <- (1 - edge.lfdr(ptmp))
  mod.gam <- cbind(mod0, uu$vectors[, 1:n.sv])
  mod0.gam <- cbind(mod0)
  ptmp <- f.pvalue(dat, mod.gam, mod0.gam)
  pprob.gam <- (1 - edge.lfdr(ptmp))
  pprob <- pprob.gam * (1 - pprob.b)
  dats <- dat * pprob
  dats <- dats - rowMeans(dats)
  uu <- eigen(t(dats) %*% dats)
  cat(paste(i, " "))
}

# this is what is normally returned by sva
# problem is that it doesnt include the diagonal matrix we want
sv_cb = svd(dats)$v[, 1:n.sv, drop = FALSE] 
svobj <- list(sv = sv_cb, pprob.gam = pprob.gam, pprob.b = pprob.b, 
              n.sv = n.sv)

# to get all info (including diagonal)
svobj_all <- svd(dats)

svs_cb <- as.data.frame(svobj$sv)

# save 
save(sv_cb, file = here("output_data", "linear_regression", "2022-05-06_prenatal_svs.RData"))

###############
#scree of SVs #
###############

# from elodie portales-casamar

# heat scree prepartion
vars <- svobj_all[['d']]^2
vars <- vars/sum(vars)
names(vars) <- paste0('SV', 1:length(vars))
xp <- 1:svobj$n.sv
varsdf <- as.data.frame(vars)
varsdf$svnum <- rep(1:length(vars))

# heat scree ggplot!! 
cb_sv_scree <- ggplot(varsdf[1:20,], aes(x=svnum, y=vars*100)) +
  geom_col() +
  theme_bw() +
  labs(x="Surrogate variable", y = "Variance (%)")

cb_sv_scree

ggsave(plot = cb_sv_scree, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal", 
                       "2022-05-06_cb_sv_heat_scree.png"),
       height = 5, width = 15, units = "cm")

# it looks like the first two svs are most important, but SVA suggests 14
# examine a correlation matrix between all svs and covariates to see
# what variance they are capturing

######################
# correlation matrix #
######################

# add svs to covariates
covariates_geno_ctpc_tech_cc <- cbind(covariates_geno_ctpc_tech_cc, sv_cb)

# rename columns with svs
colnames(covariates_geno_ctpc_tech_cc)[133:146] <-
  paste("sv",colnames(covariates_geno_ctpc_tech_cc)[133:146], sep="")

source(here("scripts", "2021-08-30_corr_mat_funs.R"))

# early exp and pn_18 are the same thing. 
# exclude early_exp from correlation matrices
corr <- corr.mat(covariates_geno_ctpc_tech_cc %>% 
                   dplyr::select(no2_preg_entire, sex, gestational_days, pss_18wk,
                                 prenatal_maternal_smoking, maternal_education_length, 
                                 cbmc.PC1, cbmc.PC2, cbmc.PC3, cbmc.PC4,
                                 genotyping.PC1, genotyping.PC2, genotyping.PC3,
                                 chip, run, row, 
                                 sv1, sv2, sv3, sv4, sv5, sv6, sv7, sv8, sv9,
                                 sv10, sv11, sv12, sv13, sv14) %>%
                   rename("Average prenatal NO2" = no2_preg_entire,
                          Sex = sex,
                          "Gestational age (days)" = gestational_days,
                          "Maternal perceived stress 18 weeks" = pss_18wk,
                          "Prenatal maternal cigarette smoking" = prenatal_maternal_smoking,
                          "Maternal education (years)" = maternal_education_length,
                          "CBMC PC1" = cbmc.PC1, "CBMC PC2" = cbmc.PC2,
                          "CBMC PC3" = cbmc.PC3, "CBMC PC4" = cbmc.PC4,
                          "Genotyping PC1" = genotyping.PC1,
                          "Genotyping PC2" = genotyping.PC2,
                          "Genotyping PC3" = genotyping.PC3,
                          Chip = chip, Run = run, Row = row, 
                          SV1 = sv1, SV2 = sv2, SV3 = sv3, SV4 = sv4, SV5 = sv5, 
                          SV6 = sv6, SV7 = sv7, SV8 = sv8, SV9 = sv9, 
                          SV10 = sv10, SV11 = sv11, SV12 = sv12, SV13 = sv13, 
                          SV14 = sv14))

pval <- pval.mat(covariates_geno_ctpc_tech_cc %>% 
                   dplyr::select(no2_preg_entire, sex, gestational_days, pss_18wk,
                                 prenatal_maternal_smoking, maternal_education_length, 
                                 cbmc.PC1, cbmc.PC2, cbmc.PC3, cbmc.PC4,
                                 genotyping.PC1, genotyping.PC2, genotyping.PC3,
                                 chip, run, row, 
                                 sv1, sv2, sv3, sv4, sv5, sv6, sv7, sv8, sv9,
                                 sv10, sv11, sv12, sv13, sv14) %>%
                   rename("Average prenatal NO2" = no2_preg_entire,
                          Sex = sex,
                          "Gestational age (days)" = gestational_days,
                          "Maternal perceived stress 18 weeks" = pss_18wk,
                          "Prenatal maternal cigarette smoking" = prenatal_maternal_smoking,
                          "Maternal education (years)" = maternal_education_length,
                          "CBMC PC1" = cbmc.PC1, "CBMC PC2" = cbmc.PC2,
                          "CBMC PC3" = cbmc.PC3, "CBMC PC4" = cbmc.PC4,
                          "Genotyping PC1" = genotyping.PC1,
                          "Genotyping PC2" = genotyping.PC2,
                          "Genotyping PC3" = genotyping.PC3,
                          Chip = chip, Run = run, Row = row, 
                          SV1 = sv1, SV2 = sv2, SV3 = sv3, SV4 = sv4, SV5 = sv5, 
                          SV6 = sv6, SV7 = sv7, SV8 = sv8, SV9 = sv9, 
                          SV10 = sv10, SV11 = sv11, SV12 = sv12, SV13 = sv13, 
                          SV14 = sv14))

svcorr <- corr.plot(pval = pval, 
                    corr = corr, 
                    label.sig = T,
                    shape = 8,shape.size = 1,
                    axis.text.size = 8, x.axis.text.angle = 90) +
  scale_y_discrete(position = "right") +
  theme(legend.position = "none",
        panel.grid = element_blank())

ggsave(plot = svcorr,
       filename = here("figures",
                       "linear_regression",
                       "prenatal", 
                       "2022-05-15_06_sv_correlations.png"),
       height = 15, width = 15, units = "cm")


######################################
# heat scree plot of svs against pcs #
######################################

source(here("scripts", "2021-09-14_heat_scree_plot.R"))

#convert raw betas to mvals
mvals <- lumi::beta2m(cb_betas_geno_cc_clean)
max(mvals)
min(mvals)

#pca 
uncor.dat <- t(scale(t(mvals)))
PCA_full<-princomp(na.omit(uncor.dat))
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance <-vars/sum(vars)

#Specify which covariates are categorical and/or continuous
meta_categorical <- 
  covariates_geno_ctpc_tech_cc[,c("sex", "prenatal_maternal_smoking",
                                  "run", "chip")] 
meta_continuous <- 
  covariates_geno_ctpc_tech_cc[,c("no2_preg_entire",
                                  "gestational_days", 
                                  "maternal_education_length", 
                                  "pss_18wk", "csed_18wk","row", "cbmc.PC1", 
                                  "cbmc.PC2", "cbmc.PC3", "cbmc.PC4", 
                                  "genotyping.PC1", "genotyping.PC2",
                                  "genotyping.PC3", "sv1", "sv2", "sv3",
                                  "sv4", "sv5", "sv6", "sv7", "sv8", "sv9", 
                                  "sv10", "sv11", "sv12", "sv13", "sv14")]

# Specify the number of PCs you want shown 
# pick enough to view most of the variance
Num<-10

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-c(1:40)

#Apply function on PCA results of uncorrected data, pulls in the meta data and beta values from above
heat_scree_plot(Loadings = Loadings, 
                Importance = Importance, 
                Num = Num, 
                Order = Order,
                Categorical = meta_categorical,
                Continuous = meta_continuous)

###########################
# add SVs to model matrix #
###########################
colnames(sv_cb) <- c(1:14)

# use 6 SVs 
modSv = cbind(mod,sv_cb)
```


Run Limma for DNAm analysis.
```{r limma}
set.seed(1234)

# fit methylation to no2
cb_fit <-  lmFit(cb_betas_geno_cc_clean, modSv)
cb_ebayes <- eBayes(cb_fit)

# subset annotation file for cpgs used in limma
annotation_cpgs <- annotation[rownames(annotation) %in% rownames(cb_ebayes),]

# check that annotation file and limma cpgs are in the same order
identical(rownames(annotation_cpgs), rownames(cb_ebayes)) # false

# reorder annotation file to match limma output
annotation_cpgs_order <- annotation_cpgs[rownames(cb_ebayes), ]

# recheck order of annotation file against limma cpgs
identical(rownames(annotation_cpgs_order), rownames(cb_ebayes)) #true

# pull out pval, coefficients from limma and combine with annotation
cb_pvals_tech <- cbind(as.data.frame(cb_ebayes$p.value), 
                       as.data.frame(cb_ebayes$coefficients),
                       as.data.frame(annotation_cpgs_order))

# rename columns to indicate pvalues vs coeffcients 
colnames(cb_pvals_tech) <- c(paste("pval", 
                                   colnames(cb_pvals_tech)[1:28], 
                                   sep = "_"), 
                             paste("coeff", 
                                   colnames(cb_pvals_tech)[29:56], 
                                   sep = "_"),
                             colnames(cb_pvals_tech)[57:89])

cb_pvals_tech$pval_no2_preg_entire_adj  <-
  p.adjust(cb_pvals_tech$pval_no2_preg_entire , method="BH")
```


Bacon pvalue adjustment
```{r bacon}
# t stats in bacon
set.seed(1234)
bc <- bacon(cb_ebayes$t[,"no2_preg_entire"])

# get bacon pvals
cb_pvals_tech$pval_no2_preg_entire_bacon <- pval(bc)

# bh adjusted bacon pvals
cb_pvals_tech$pval_no2_preg_entire_bacon_adj  <-
  p.adjust(cb_pvals_tech$pval_no2_preg_entire_bacon , method="BH")

bacon_before_after <- plot(bc, type="qq") +
  scale_color_manual(values = c("#254A90")) +
  theme_bw() + 
  theme(legend.position = "none")

ggsave(plot = bacon_before_after, 
       filename= here("figures", 
                      "linear_regression", 
                      "prenatal", 
                      "2022-05-06_cb_bacon_before_after.png"),
       width = 12, height = 6, units = "cm")
```


Examine the qqplot.
```{r qqplot_pval_histo}

# use code from qqman::qq function 
# plot with ggplot to make it more aesthetically appealling 

###############
# raw pvalues #
###############

pvector <- cb_pvals_tech$pval_no2_preg_entire

pvector2 <- pvector[!is.na(pvector) & !is.nan(pvector) & !is.null(pvector) & 
                      is.finite(pvector) & pvector < 1 & pvector > 0]

pval_qq <- as.data.frame(cbind(o = -log10(sort(pvector2, decreasing = FALSE)),
                               e = -log10(ppoints(length(pvector2)))))

# inflation and bias
inflation <- inflation(bc)
bias <-  bias(bc)

# inflation and bias after bacon correction
inflation_after <- inflation(bacon(tstat(bc)))
bias_after <- bias(bacon(tstat(bc)))

# qqplot of bacon corrected pvalues for main paper figure
before_qqplot <-  ggplot(pval_qq, aes(x=e, y=o)) + 
  geom_point(col="#254A90", size=1) + 
  geom_abline(slope = 1, intercept = 0, 
              col = "red", size=0.25, linetype="dashed") +
  labs(x = "Expected -log10(p-value)",
       y = "Observed -log10(p-value)") +
  annotate("text", x=3, y=1.5, 
           label = paste("Inflation = ", signif(inflation, 2), 
                         "\nBias = ", signif(bias, 2)), 
           size=3, hjust=0) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(size=9),
        axis.text.y = element_text(size=9),
        axis.title = element_blank()) +
  facet_grid(.~"Uncorrected") 


ggsave(plot = before_qqplot, 
       filename= here("figures", 
                      "linear_regression", 
                      "prenatal", 
                      "2022-05-06_uncorrected_qqplot.png"),
       width = 6, height = 5, units = "cm")

###########################
# bacon corrected pvalues #
###########################

pvector <- cb_pvals_tech$pval_no2_preg_entire_bacon

pvector2 <- pvector[!is.na(pvector) & !is.nan(pvector) & !is.null(pvector) & 
                      is.finite(pvector) & pvector < 1 & pvector > 0]

pval_qq <- as.data.frame(cbind(o = -log10(sort(pvector2, decreasing = FALSE)),
                               e = -log10(ppoints(length(pvector2)))))

# inflation and bias after bacon correction
set.seed(1234)
inflation_after <- inflation(bacon(tstat(bc)))
bias_after <- bias(bacon(tstat(bc)))

# inflation and bias after bacon correction
inflation_after <- inflation(bacon(tstat(bc)))
bias_after <- bias(bacon(tstat(bc)))

# qqplot of bacon corrected pvalues for main paper figure
after_qqplot <-  ggplot(pval_qq, aes(x=e, y=o)) + 
  geom_point(col="#254A90", size=1) + 
  geom_abline(slope = 1, intercept = 0, 
              col = "red", size=0.25, linetype="dashed") +
  labs(x = "Expected -log10(p-value)",
       y = "Observed -log10(p-value)") +
  annotate("text", x=3, y=1.5, 
           label = paste("Inflation = ", signif(inflation_after, 2), 
                         "\nBias = ", signif(bias_after, 2)), 
           size=3, hjust=0) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(size=9),
        axis.text.y = element_text(size=9),
        axis.title = element_blank()) +
  facet_grid(.~"Bacon corrected") +
  lims(y = c(0, 7))

ggsave(plot = after_qqplot, 
       filename= here("figures", 
                      "linear_regression", 
                      "prenatal", 
                      "2022-05-06_bacon_corrected_qqplot.png"),
       width = 6, height = 5, units = "cm")
```


Delta beta
```{r db}
range = max(covariates_geno_ctpc_tech_cc$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc$no2_preg_entire)

cb_pvals_tech$db <- cb_pvals_tech$coeff_no2_preg_entire * range
```


Volcano plot
```{r volcano_db_tech}

#make new column to label sig points with cpgid
cb_pvals_tech$cpgid <- rownames(cb_pvals_tech)

# plot that volcano
cb_volcano <- ggplot(cb_pvals_tech, 
                     aes(x=as.numeric(db) , 
                         y=-log10(as.numeric(pval_no2_preg_entire_bacon)))) +
  geom_point(alpha=0.4, col = "#254A90", size=1) +
  # colour significant cpgs; 2.5% effect size cut off and FDR < 0.25
  geom_point(data=subset(cb_pvals_tech, 
                         (pval_no2_preg_entire_bacon_adj < 0.25 & db > 0.025) | 
                           (pval_no2_preg_entire_bacon_adj < 0.25 & db < -0.025)), 
             alpha=0.4, col = "red", size=1) + 
  # horizontal lines for effect size cut offs
  #geom_vline(xintercept = c(2.5, -2.5), linetype = "dashed", size=0.25) +
  # aesthetics
  scale_x_continuous(breaks = seq(-0.50, 0.50, by = 0.10)) +
  labs(y="-log10(unadjusted p-value)", x = "Effect size (% DNAm)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.text.x = element_text(size=9),
        axis.text.y = element_text(size=9),
        axis.title = element_blank()) 
cb_volcano

ggsave(plot = cb_volcano,
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal", 
                       "2022-05-10_cb_volcanoplot.png"),
       width = 7.5, height = 6, units = "cm")
```


Subset out important cpgs
```{r sig_cpgs_tables}

# add effect size to ebayes
cb_ebayes$db <- cb_pvals_tech$db

# subset based on effect size
cb_ebayes_sig <- cb_ebayes[cb_ebayes$db > 0.025 | cb_ebayes$db < -0.025, ]

# get top ten significant genes with coefficients and write csv
write.csv(limma::topTable(cb_ebayes_sig, number=20, 
                          coef ="no2_preg_entire", sort.by = "P", confint = T),
          here("tables", "prenatal", "2022-05-06_toptable_prenatal_10cpgs.csv"))

# get top 10 cpgs with manifest annotation
write.csv(subset(cb_pvals_tech, 
                 (pval_no2_preg_entire_bacon < 0.0000195 & db > 0.025) | 
                   (pval_no2_preg_entire_bacon < 0.0000195 & db < -0.025)) %>%
            arrange(pval_no2_preg_entire_bacon), 
          here("tables", "prenatal", 
               "2022-05-06_top10_prenatal_cpgs.csv"))

# write out all prenatal findings
write.csv(cb_pvals_tech, here("tables", "prenatal", 
                              "2022-05-06_ALL_prenatal_cpgs.csv"))
```


Plot FDR significant CpGs from volcano plot.
```{r sig_cpgs_plots}

###############
# cg27481559  #
###############

#add dnam to covariates for linear regression
covariates_geno_ctpc_tech_cc$cb_cg27481559 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg27481559", ]

#linear regression
cb_cg27481559_lm <- lm(cb_cg27481559 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 +
                         sv10 + sv11 + sv12 + sv13 + sv14,
                       data = covariates_geno_ctpc_tech_cc)

# get summary of  linear regression
summary(cb_cg27481559_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
prenatal_no2_me <- ggpredict(cb_cg27481559_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("cb_cg27481559" = predicted,
         "no2_preg_entire" = x)

# plot
cb_cg27481559_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                             aes(x = no2_preg_entire, 
                                 y= cb_cg27481559)) + 
  geom_line(data = prenatal_no2_me, 
            aes(x=no2_preg_entire, y = cb_cg27481559),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = prenatal_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  scale_y_continuous(limits = c(0.15,0.45), 
                     breaks=seq(0.20,0.40,0.10),
                     labels = scales::number_format(accuracy = 0.01)) +
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = cb_cg27481559_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal", 
                       "2022-05-10_cb_cg27481559.png"),
       width = 3.7, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc$cb_cg27481559, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") #-0.4021566

#############################
# AFAP1/LOC84740 cg12228123 #
#############################

# add dnam to covariates for linear regression
covariates_geno_ctpc_tech_cc$cb_cg12228123<- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg12228123", ]

# linear regression
cb_cg12228123_lm <- lm(cb_cg12228123 ~ no2_preg_entire + 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 +
                         sv10 + sv11 + sv12 + sv13 + sv14,
                       data=covariates_geno_ctpc_tech_cc)

# get summary of fit
summary(cb_cg12228123_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
cb_cg12228123_no2_me <- ggpredict(cb_cg12228123_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("cb_cg12228123" = predicted,
         "no2_preg_entire" = x)

# plot
cb_cg12228123_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                             aes(x = no2_preg_entire, 
                                 y= cb_cg12228123)) + 
  geom_line(data = cb_cg12228123_no2_me, 
            aes(x=no2_preg_entire, y = cb_cg12228123),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data= cb_cg12228123_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  scale_y_continuous(limits = c(0.50,1.0), breaks=seq(0.55,0.95,0.20)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = cb_cg12228123_plot,
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_cb_cg12228123_AFAP1.png"),
       width = 3.7, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc$cb_cg12228123, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") # -0.4331415

```


### Examine genome wide methylation changes vs no2 exposure

Examine mean methylation vs no2
```{r mean_meth}
# convert betas to df
cb_betas_geno_cc_clean_df <- as.data.frame(cb_betas_geno_cc_clean)

methmeans <- colMeans(cb_betas_geno_cc_clean_df)

covariates_geno_ctpc_tech_cc <- cbind(covariates_geno_ctpc_tech_cc, methmeans)

#colmeans
meanmeth_lm <- 
  lm(methmeans ~ no2_preg_entire + sex + gestational_days + pss_18wk + 
       prenatal_maternal_smoking + maternal_education_length + 
       cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 +
       genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
       sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 + sv10 +
       sv11 + sv12 + sv13 + sv14, 
     data=covariates_geno_ctpc_tech_cc)

#get summary
summary(meanmeth_lm) #no2 not sig

# estimate marginal effects of prenatal NO2 exposure on DNAm
globaldna_no2_me <- ggpredict(meanmeth_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("methmeans" = predicted,
         "no2_preg_entire" = x)

# plot 
global_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                           aes(x = no2_preg_entire, 
                               y= methmeans)) + 
  geom_line(data = globaldna_no2_me, 
            aes(x=no2_preg_entire, y = methmeans),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = globaldna_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) +
  lims(y=c(0.49,0.54))

ggsave(plot = global_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_cb_global_dnam.png"),
       width = 4.0, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc$methmeans, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") # 0.09068486

# effect size
range_cb = max(covariates_geno_ctpc_tech_cc$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc$no2_preg_entire)

summary(meanmeth_lm)$coefficients["no2_preg_entire", "Estimate"]*range_cb # -0.1103026
```


### Examine LINE methylation vs NO2

```{r line_dnam}
# read in line annotation from UCSC and repeat masker 
repeats_450k <- read.delim(file = gzfile(here("input_data", 
                                              "ucsc_rmsk_lines", 
                                              "ucsc_repeatmasker_enc_methyl450_gm12878.gz")),
                           header = T, skip = 1, sep = "\t")

# subset repeats for L1 LINE elements that are not in sex chromosomes
lines_450k  <- repeats_450k %>% 
  subset(rmsk.repClass == "LINE") %>%
  subset(rmsk.repFamily == "L1") %>%
  subset(!wgEncodeHaibMethyl450Gm12878SitesRep1.chrom %in% c("chrX", "chrY")) %>%
  subset(!is.na(wgEncodeHaibMethyl450Gm12878SitesRep1.chromStart))


# subset betas for those that overlap with line elements
cb_betas_geno_cc_clean_df_lines <- cb_betas_geno_cc_clean_df %>% 
  subset(rownames(.) %in% lines_450k$wgEncodeHaibMethyl450Gm12878SitesRep1.name)

# num of cpgs annotated to lines 
nrow(cb_betas_geno_cc_clean_df_lines) # 7302 

# get mean of line meth 
line_meth <- colMeans(cb_betas_geno_cc_clean_df_lines)

# cbind line methylation to covariate data frame
covariates_geno_ctpc_tech_cc <- cbind(covariates_geno_ctpc_tech_cc, line_meth)

#colmeans
line_meth_lm <- lm(line_meth ~ no2_preg_entire + 
                     sex + gestational_days + pss_18wk +   
                     prenatal_maternal_smoking + maternal_education_length + 
                     cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                     genotyping.PC1 + genotyping.PC2 + genotyping.PC3 +
                     sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 + sv10 +
                     sv11 + sv12 + sv13 + sv14, 
                   data=covariates_geno_ctpc_tech_cc)

# summary of line regression
summary(line_meth_lm) #no2 not sig


# estimate marginal effects of prenatal NO2 exposure on DNAm
line_meth_no2_me <- ggpredict(line_meth_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("line_meth" = predicted,
         "no2_preg_entire" = x)

# plot 
line_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                         aes(x = no2_preg_entire, 
                             y= line_meth)) + 
  geom_line(data = line_meth_no2_me, 
            aes(x=no2_preg_entire, y = line_meth),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = line_meth_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) +
  lims(y=c(0.81,0.86))

ggsave(plot = line_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_cb_line_dnam.png"),
       width = 4.0, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc$line_meth, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") # 0.1115208

# effect size
summary(line_meth_lm)$coefficients["no2_preg_entire", "Estimate"]*range_cb # -0.1103026
```


### Gene specific changes

These genes are related to the proposed pathway of air pollution action or have previously been shown to exhibit DNA methylation changes associated with prenatal air pollution exposure. 

```{r gene_specific}
#FHC is FTH1 (ferritin heavy chain 1)
#nothing maps to FOXP3
#TNFA is TNF
goi <- annotation[grep("^RNF39|^CYP2E1|^LONP1|^SLC25A28|^PLVAP|^CAT$|^CAT;|^TPO|^NFKB1|^NOX2|^SOD1|^SOD2|^TRX1|^TRX2|^GSTP1|^MT3|^NQO1|^HMOX1|^TGPX1|^DDH1|^NOS2|^CYP2C11|^CYP7B|^ALOX12$|^ALOX5;|^ALOX5$|^NRF2$|^NRF2;|^KEAP1|^AHR$|^AHR;|^ARNT;|^ARNT$|^GCLC|^GCLM|^GSR1|^SLC7A11|^GSTA1|^GSTA2|^GSTA3|^GSTA5|^GSTM1|^GSTM2|^GSTM3|^TXN$|^TXN;|^TXNRD1$|^TXNRD1;|^SRXN1|^G6PD|^CYP1A1|^IL1A|^IL2$|^IL2;|^IL3$|^IL3;|^IL4;|^IL4$|^IL5$|^IL5;|^IL6$|^IL6;|^IL9$|^IL9;|^IL10$|^IL10;|^IL12A|^IL13|^IL15$|^IL15;|^IL17A|^IL18$|^IL18;|^IL19|^IL21$|^IL21;|^IL22$|^IL22;|^IL23A$|^IL23A;|^IL25$|^IL25;|^IL27$|^IL27;|^IL31$|^IL31;|^IL33$|^IL33;|^IFNG$|^IFNG;|^TNF$|^TNF;|^TGFB1$|^TGFB1;|^FOXP3|^FTH1", annotation$UCSC_RefGene_Name), ] %>% as.data.frame()


unique(goi$UCSC_RefGene_Name)
# only first annotate gene
unique(gsub(";.*$", "", unique(goi$UCSC_RefGene_Name))) # 59


# transpose for lm function
cb_betas_geno_cc_clean_t <- t(cb_betas_geno_cc_clean)

#apply function for calculating significance of associations using mice and imputed covariate data
prenatal_goi_list <- 
  pbapply(cb_betas_geno_cc_clean_t[,colnames(cb_betas_geno_cc_clean_t ) %in% rownames(goi)], 2, function(x) {
    
    combined <- cbind(covariates_geno_ctpc_tech_cc, x)
    
    #run the linear regression on this cpg using all 5 imputed covariate data sets
    fit <- lm(x ~ no2_preg_entire + 
                sex + gestational_days + pss_18wk +  
                prenatal_maternal_smoking + maternal_education_length + 
                cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 + sv10 +
                sv11 + sv12 + sv13 + sv14, 
              data=combined)
    
    #get summary of lm fit
    #this includes coefficients and pvalues
    #coerce into df for manipulation
    #add in terms (rownames) as column for annotation
    fit_sum <- summary(fit)
    fit_sum_coeff <- as.data.frame(fit_sum$coefficients)
    fit_sum_coeff$term <- rownames(fit_sum_coeff)
    
    #this data frame has rows for each linear regression variable, and columns for statistics
    #want a dataframe of 1 row and many columns (one for each variable with each statistic type)
    #do this by using the spread function for each statistic
    #add cpg id onto dataframe for pvalues (to keep track of information)
    #then cbind dataframes together
    
    #coefficient
    estimate <- spread(fit_sum_coeff[,c("term", "Estimate")], term, Estimate)
    colnames(estimate) <- paste("estimate", colnames(estimate), sep=".")
    
    #std error
    std.error <- spread(fit_sum_coeff[,c("term", "Std. Error")], term, 'Std. Error')
    colnames(std.error) <- paste("std.error", colnames(std.error), sep=".")
    
    #t statistics
    statistic <- spread(fit_sum_coeff[,c("term", "t value")], term, 't value')
    colnames(statistic) <- paste("t.val", colnames(statistic), sep=".")
    
    #pvalue
    p.value <- spread(fit_sum_coeff[,c("term", "Pr(>|t|)")], term, 'Pr(>|t|)')
    colnames(p.value) <- paste("p.value", colnames(p.value), sep=".")
    
    #add in cpgid for annotation 
    p.value$cpg <- colnames(x)
    
    #cbind into wide dataframe
    widepool <- cbind(estimate, std.error, statistic, p.value)
    
    #return the wide dataframe
    #these will be do.call(rbind()) together after
    return(widepool)
    
  })


prenatal_goi <- do.call(rbind, prenatal_goi_list)

prenatal_goi$p.value.no2_preg_entire_adjusted <- 
  p.adjust(prenatal_goi$p.value.no2_preg_entire, method="BH")

#reorder according ot unadjusted no2 pvalue
prenatal_goi_order <- prenatal_goi[order(prenatal_goi$p.value.no2_preg_entire),]

#reorder goi annotation to be in same order as goi stats
goi_order <- goi[rownames(prenatal_goi_order),]

#check identical
identical(rownames(goi_order), rownames(prenatal_goi_order)) #true

#bind
goi_df <- cbind(prenatal_goi_order, goi_order)

#sort goi df by pvalue
goi_df_pval <- goi_df %>% arrange(p.value.no2_preg_entire)

# add delta beta
goi_df_pval_db <- merge(goi_df_pval, cb_pvals_tech, by="row.names", all.x=T) 

# write out table for all GOIs
write.csv(goi_df_pval_db, file=here("tables", "prenatal",
                                    "2022-05-06_goi_prenatal_cpgs.csv"))

# write table of top ten 
write.csv(goi_df_pval_db %>% 
            subset(db > 0.025 | db < -0.025) %>% 
            subset(-log10(p.value.no2_preg_entire) > 1.45) %>%
            arrange(p.value.no2_preg_entire),
          here("tables", "prenatal", "2022-05-06_top10goi_prenatal_cpgs.csv"))

# use top table to get 95% CI
goi_df_pval_db_sig <- goi_df_pval_db %>% 
  subset(db > 0.025 | db < -0.025) %>% 
  subset(-log10(p.value.no2_preg_entire) > 1.45)

write.csv(goi_df_pval_db_sig, here("tables", "prenatal", "2022-05-06_toptable_prenatal_goi.csv"))

#check cpgs from gruvieva 2017 paper

##################
#LONP1 cg12283362#
##################

covariates_geno_ctpc_tech_cc$cg12283362 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg12283362", ] 

cg12283362_lm <- lm(cg12283362 ~ 
                      no2_preg_entire + sex + gestational_days + pss_18wk + 
                      prenatal_maternal_smoking + maternal_education_length + 
                      cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                      genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                      sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 + 
                      sv10 + sv11 + sv12 + sv13 + sv14, 
                    data=covariates_geno_ctpc_tech_cc)

# get lm info
summary(cg12283362_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
cg12283362_no2_me <- ggpredict(cg12283362_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("cg12283362" = predicted,
         "no2_preg_entire" = x)

# plot 
cb_cg12283362_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                                  aes(x = no2_preg_entire, 
                                      y= cg12283362)) + 
  geom_line(data = cg12283362_no2_me, 
            aes(x=no2_preg_entire, y = cg12283362),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = cg12283362_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
    scale_y_continuous(breaks = seq(0.6,0.9,by=0.1), 
                       limits = c(0.6,0.9),
                       labels = scales::number_format(accuracy = 0.01)) +

  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = cb_cg12283362_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_cb_cg12283362_dnam.png"),
       width = 4.0, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc$cg12283362, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") # -0.03698505

# effect size
range_cb = max(covariates_geno_ctpc_tech_cc$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc$no2_preg_entire)

summary(cg12283362_lm)$coefficients["no2_preg_entire", "Estimate"]*range_cb # -0.1103026


#####################
#SLC25A28 cg08973675#
#####################

covariates_geno_ctpc_tech_cc$cg08973675 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg08973675", ] 

cg08973675_lm <- lm(cg08973675~ 
                      no2_preg_entire + sex + gestational_days + pss_18wk + 
                      prenatal_maternal_smoking + maternal_education_length + 
                      cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4 + 
                      genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                      sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 + 
                      sv10 + sv11 + sv12 + sv13 + sv14, 
                    data=covariates_geno_ctpc_tech_cc)

summary(cg08973675_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
cg08973675_no2_me <- ggpredict(cg08973675_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("cg08973675" = predicted,
         "no2_preg_entire" = x)

# plot 
cb_cg08973675_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc,
                                  aes(x = no2_preg_entire, 
                                      y= cg08973675)) + 
  geom_line(data = cg08973675_no2_me, 
            aes(x=no2_preg_entire, y = cg08973675),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = cg08973675_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  scale_y_continuous(breaks = seq(0,0.16,by=0.04), limits = c(0,0.16)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = cb_cg08973675_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_cb_cg08973675_dnam.png"),
       width = 4.0, height = 3.2, units = "cm")


# correlation 
cor(covariates_geno_ctpc_tech_cc$cg08973675, 
    covariates_geno_ctpc_tech_cc$no2_preg_entire,
    method="pearson") # 0.1007488

# effect size
range_cb = max(covariates_geno_ctpc_tech_cc$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc$no2_preg_entire)

summary(cg08973675_lm)$coefficients["no2_preg_entire", "Estimate"]*range_cb # -0.1103026
```



Examine whether DNAm at significant cpgs is associated with reported wheeze at age one. In previous CHILD paper [(PMID: 26424567)](https://www.science.org/doi/10.1126/scitranslmed.aab2271) wheeze was defined using wheeze questionnaires. If the child had wheezed with or without a cold during the first year of life (recorded via questionnaires answered by parents at 3, 6, and 12 months), the child was included in the wheezing group. Children were also included in the wheezing group if the CHILD clinician recorded a wheeze during the 1-year clinical assessment.
```{r prenatal_cpgs_age_one_wheeze}

covariates_geno_ctpc_tech_cc$any_wheeze_y1 <- 
  ifelse(covariates_geno_ctpc_tech_cc$reported_wheezing_y1 ==1 | 
           covariates_geno_ctpc_tech_cc$clinician_wheeze_dx_y1 ==1, 1, 0)

#############################
# AFAP1/LOC84740 cg12228123 #
#############################

# reported wheeze
ggplot(covariates_geno_ctpc_tech_cc %>% 
         dplyr::select(any_wheeze_y1, cb_cg12228123) %>%
         na.omit(), 
       aes(x=as.factor(any_wheeze_y1), y=cb_cg12228123)) +
  geom_boxplot() +
  geom_jitter(height=0) +
  theme_bw()

# wheeze logistic
cg12228123_wheeze_log <- glm(any_wheeze_y1 ~ cb_cg12228123 +
                               sex + gestational_days + pss_18wk +
                               prenatal_maternal_smoking + maternal_education_length + 
                               cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                               genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                               sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 +
                               sv10 + sv11 + sv12 + sv13 + sv14,                          data=covariates_geno_ctpc_tech_cc[!is.na(covariates_geno_ctpc_tech_cc$any_wheeze_y1),],
                             family = "binomial")

summary(cg12228123_wheeze_log)
cg12228123_wheeze_log_tidy <- tidy(cg12228123_wheeze_log, exponentiate = T, conf.int = T) 
cg12228123_wheeze_log_tidy


# clinician atopy
ggplot(covariates_geno_ctpc_tech_cc %>% 
         dplyr::select(clinician_atopy_dx_y1, cb_cg12228123) %>%
         na.omit(), 
       aes(x=clinician_atopy_dx_y1, y=cb_cg12228123)) +
  geom_boxplot() +
  geom_jitter(height=0) +
  theme_bw()

# atopy logistic
cg12228123_atopy_log <- glm(clinician_atopy_dx_y1 ~ cb_cg12228123 +
                              sex + gestational_days + pss_18wk +
                              prenatal_maternal_smoking + maternal_education_length + 
                              cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                              genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                              sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + sv9 +
                              sv10 + sv11 + sv12 + sv13 + sv14,                          data=covariates_geno_ctpc_tech_cc[!is.na(covariates_geno_ctpc_tech_cc$any_wheeze_y1),],
                            family = "binomial")

summary(cg12228123_atopy_log)
cg12228123_atopy_log_tidy <- tidy(cg12228123_atopy_log, exponentiate = T, conf.int = T) 
cg12228123_atopy_log_tidy

###############
# cg27481559  #
###############

# wheeze
ggplot(covariates_geno_ctpc_tech_cc %>% 
         dplyr::select(any_wheeze_y1, cb_cg27481559) %>%
         na.omit(), 
       aes(x=as.factor(any_wheeze_y1), y=cb_cg27481559)) +
  geom_boxplot() +
  geom_jitter(height=0) +
  theme_bw()

# wheeze logistic 
cb_cg27481559_log <- glm(any_wheeze_y1 ~ 
                           cb_cg27481559 +  
                           sex + gestational_days + pss_18wk +
                           prenatal_maternal_smoking + 
                           maternal_education_length + 
                           cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                           genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                           sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                           sv9 + sv10 + sv11 + sv12 + sv13 + sv14,                          data=covariates_geno_ctpc_tech_cc[!is.na(covariates_geno_ctpc_tech_cc$any_wheeze_y1),],
                         family = "binomial")


summary(cb_cg27481559_log)

cb_cg27481559_log_tidy <- tidy(cb_cg27481559_log, exponentiate = T, conf.int = T) 
cb_cg27481559_log_tidy

# atopy
ggplot(covariates_geno_ctpc_tech_cc %>% 
         dplyr::select(clinician_atopy_dx_y1, cb_cg27481559) %>%
         na.omit(), 
       aes(x=as.factor(clinician_atopy_dx_y1), y=cb_cg27481559)) +
  geom_boxplot() +
  geom_jitter(height=0) +
  theme_bw()

# wheeze logistic 
cb_cg27481559_atopy_log <- glm(clinician_atopy_dx_y1 ~ 
                                 cb_cg27481559 +  
                                 sex + gestational_days + pss_18wk +
                                 prenatal_maternal_smoking + 
                                 maternal_education_length + 
                                 cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                                 genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                                 sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                                 sv9 + sv10 + sv11 + sv12 + sv13 + sv14,                          data=covariates_geno_ctpc_tech_cc[!is.na(covariates_geno_ctpc_tech_cc$any_wheeze_y1),],
                               family = "binomial")


summary(cb_cg27481559_atopy_log)

cb_cg27481559_atopy_log_tidy <- tidy(cb_cg27481559_atopy_log, exponentiate = T, conf.int = T) 
cb_cg27481559_atopy_log_tidy
```



Try re-running limma with birth month/ interaction between birth month and air pollution.
```{r interaction_terms_to_assess_birthmonth}
#want to use anova to compare models usign anova
#anova only accepts models from lm
#use lm to compare how birth month affects cpgs above 10^-3 sig 

#You should be able to ask for confidence intervals for the slope coefficients for each month.  #even in a tabular display, you may see that smoothly varying pattern we were discussing.
#Then, you could use those to plot a sort of heres how the slope varies by month picture.Sort #of like forest plots for meta analysis results, if that makes sense.

#Also, before we decide it doesnt matter, the individual slopes might look fairly similar, but the appropriate overall test would be to check the anova(fit_with_interaction) and check the p-value on the interaction term.

#start by investigating cpgs identified as nominally significant in volcano plot

###################
#cg12228123 AFAP1 #             
###################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg12228123 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg12228123", ] 

# fit wiht lm
cg12228123_bm_lm <- lm(cb_cg12228123 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg12228123_bm_lm)

# get plot of birth month effect with 95% CI
cg12228123_bm_plot <- tidy(cg12228123_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg12228123_AFAP1_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


##############
# cg27481559 #             
##############

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg27481559 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg27481559", ] 

# fit wiht lm
cg27481559_bm_lm <- lm(cb_cg27481559 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg27481559_bm_lm)

# get plot of birth month effect with 95% CI
cg27481559_bm_plot <- tidy(cg27481559_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg27481559_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


###################
# cg22230559 BAT1 #             
###################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg22230559 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg22230559", ] 

# fit wiht lm
cg22230559_bm_lm <- lm(cb_cg22230559 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg22230559_bm_lm)

# get plot of birth month effect with 95% CI
cg22230559_bm_plot <- tidy(cg22230559_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg22230559_BAT1_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


##############
# cg05231970 #             
##############

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg05231970 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg05231970", ] 

# fit wiht lm
cg05231970_bm_lm <- lm(cb_cg05231970 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg05231970_bm_lm)

# get plot of birth month effect with 95% CI
cg22230559_bm_plot <- tidy(cg05231970_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg05231970_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


###################
# cg26787539 CDT1 #             
###################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg26787539 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg26787539", ] 

# fit wiht lm
cg26787539_bm_lm <- lm(cb_cg26787539 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg26787539_bm_lm)

# get plot of birth month effect with 95% CI
cg26787539_bm_plot <- tidy(cg26787539_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg26787539_CDT1_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


##############
# cg08568767 #             
##############

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg08568767 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg08568767", ] 

# fit wiht lm
cg08568767_bm_lm <- lm(cb_cg08568767 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg08568767_bm_lm)

# get plot of birth month effect with 95% CI
cg08568767_bm_plot <- tidy(cg08568767_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg08568767_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


#######################
# cg02386875 LRRC37B2 #             
#######################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg02386875 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg02386875", ] 

# fit wiht lm
cg02386875_bm_lm <- lm(cb_cg02386875 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg08568767_bm_lm)

# get plot of birth month effect with 95% CI
cg02386875_bm_plot <- tidy(cg02386875_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg02386875_LRRC37B2_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


####################
# cg04256907 ZMIZ1 #             
####################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg04256907 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg04256907", ] 

# fit wiht lm
cg04256907_bm_lm <- lm(cb_cg04256907 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg04256907_bm_lm)

# get plot of birth month effect with 95% CI
cg04256907_bm_plot <- tidy(cg04256907_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg04256907_ZMIZ1_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


###################
# cg10920410 GMPS #             
###################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg10920410 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg10920410", ] 

# fit wiht lm
cg10920410_bm_lm <- lm(cb_cg10920410 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg10920410_bm_lm)

# get plot of birth month effect with 95% CI
cg10920410_bm_plot <- tidy(cg10920410_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg10920410_GMPS_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")


#######################
# cg03392571 C21orf70 #             
#######################

# add to covariate data frame
covariates_geno_ctpc_tech_cc$cb_cg03392571 <- 
  cb_betas_geno_cc_clean[rownames(cb_betas_geno_cc_clean)=="cg03392571", ] 

# fit wiht lm
cg03392571_bm_lm <- lm(cb_cg03392571 ~ 
                         no2_preg_entire + sex + gestational_days + pss_18wk +
                         prenatal_maternal_smoking + maternal_education_length + 
                         cbmc.PC1 + cbmc.PC2 + cbmc.PC3 + cbmc.PC4  + 
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 + 
                         sv1 + sv2 + sv3 + sv4 + sv5 + sv6 + sv7 + sv8 + 
                         sv9 + sv10 + sv11 + sv12 + sv13 + sv14 + 
                         birth_month + no2_preg_entire*birth_month,
                       data=covariates_geno_ctpc_tech_cc)

# anova
anova(cg03392571_bm_lm)

# get plot of birth month effect with 95% CI
cg03392571_bm_plot <- tidy(cg03392571_bm_lm, conf.int = 0.95) %>% 
  filter(term %>% str_detect(":")) %>% 
  mutate(term = factor(term, levels=c("no2_preg_entire:birth_monthfebruary",
                                      "no2_preg_entire:birth_monthmarch",
                                      "no2_preg_entire:birth_monthapril",
                                      "no2_preg_entire:birth_monthmay",
                                      "no2_preg_entire:birth_monthjune",
                                      "no2_preg_entire:birth_monthjuly",
                                      "no2_preg_entire:birth_monthaugust",
                                      "no2_preg_entire:birth_monthseptember",
                                      "no2_preg_entire:birth_monthoctober",
                                      "no2_preg_entire:birth_monthnovember",
                                      "no2_preg_entire:birth_monthdecember"))) %>% 
  ggplot(aes(y = estimate*100, x = as.factor(term))) + 
  geom_point() +
  geom_errorbar(aes(ymin = conf.low*100 , ymax = conf.high*100)) +
  scale_x_discrete(labels=c("February","March", "April", "May",
                            "June","July", "August", "September",
                            "October", "November","December")) +
  ylim(-30, 30) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90),
        axis.text = element_text(size=9, hjust=0.9),
        axis.title = element_blank()) 

ggsave(filename="2022-05-06_CHILD_cb_cg03392571_C21orf70_bm.png",
       path =  here("figures", "linear_regression", "prenatal"),
       width = 4.8, height = 4, units = "cm")
```



"Table 1" of prenatal model
```{r prenatal_table1}
library(table1)

# rename rownames of cbmc estimates
identical(rownames(child_fscbc_ecc2_cb_df_nodups), 
          rownames(cb_pdata))
child_fscbc_ecc2_cb_df_nodups$sampleid <- cb_pdata$Sample_Label

# bind cell type estimates with covariate info
covariates_geno_ctpc_tech_cc_cbmc <- merge(covariates_geno_ctpc_tech_cc, 
                                           child_fscbc_ecc2_cb_df_nodups, 
                                           by = "sampleid")

# table 1
table1(~sex + gestational_days + pss_18wk + no2_preg_entire + 
         maternal_education_length +  maternal_ethnicity + 
         prenatal_maternal_smoking + CD4T + CD8T + nRBC + Bcell + Mono + NK,
       data = covariates_geno_ctpc_tech_cc_cbmc)

# prenatal no2 quartiles
quantile(covariates_geno_ctpc_tech_cc_cbmc$no2_preg_entire)
```


#### Examine persistence of prenatal changes at age one

First we need to prepare the age one data for the linear model. Then we need to calculate the number of SVs needed for this model. 

Get age one pdata
```{r y1_pdata}
#subset out age one pdata
y1_pdata <- subset(pdata_nodups, Tissue == "P")

#check size
dim(y1_pdata) # 145 17
```


Get age one betas
```{r y1_betas}
#subset out PBMC betas 
y1_betas <- betas[ ,colnames(betas) %in% rownames(y1_pdata)]

#check if age one betas and pdata are in same order
identical(colnames(y1_betas), rownames(y1_pdata)) # TRUE

#rename colums with sample label (5 dig num)
colnames(y1_betas) <- y1_pdata$Sample_Label

dim(y1_betas) # 145
```


Load age one cell type PCs
```{r y1_celltypePCs}
#load age one PBMC estimates
load(here("output_data", "deconvolution", "2021-09-24_PBMC_PCs.RData"))

#rename columns of PBMCs for clarity
colnames(y1_pcs) <- paste("pbmc", colnames(y1_pcs), sep=".")

#rename row of cell type PCs to sample label (5 digit num)
identical(rownames(y1_pcs), rownames(y1_pdata)) #true
rownames(y1_pcs) <- y1_pdata$Sample_Label

#subset cell type PCs for complete cases
y1_pcs_sub <- y1_pcs[rownames(y1_pcs) %in% covariates_geno_ctpc_tech_cc$sampleid,]
dim(y1_pcs_sub) # 95

# add age one PBMCs to covariate data frame
covariates_geno_ctpc_tech_cc_y1 <- cbind(covariates_geno_ctpc_tech_cc, y1_pcs_sub)
```


Add age one batch variables to covariate data frame
```{r y1_data_prep}
#add age one batch variables to covariate matrix
y1_pdata$y1_run <- as.factor(y1_pdata$Run) 
y1_pdata$y1_row  <- as.factor(substr(y1_pdata$Sentrix_Position,1,3)) 
y1_pdata$y1_row <- as.numeric(y1_pdata$y1_row)
y1_pdata$y1_chip <- as.factor(y1_pdata$Sentrix_ID)

covariates_geno_ctpc_tech_cc_y1 <- cbind(covariates_geno_ctpc_tech_cc_y1, 
                                         y1_pdata[y1_pdata$Sample_Label %in% covariates_geno_ctpc_tech_cc_y1$sampleid, c("y1_run", "y1_row", "y1_chip")])
```


Subset for complete cases in age one data
```{r y1_complete_cases}
#subset covariates with age one cell type PCs for linear regression
covariates_geno_ctpc_tech_cc_y1_sub <- covariates_geno_ctpc_tech_cc_y1 %>%
  subset(complete.cases(no2_preg_entire, sex, maternal_smoking_y1,
                        maternal_education_length, pss_6mos,
                        pbmc.PC1, pbmc.PC2, pbmc.PC3,
                        genotyping.PC1, genotyping.PC2, genotyping.PC3))

#check size
dim(covariates_geno_ctpc_tech_cc_y1_sub) # 92

#subset age one betas for samples in the complete case cord blood betas 
y1_betas_cc <- y1_betas[,colnames(y1_betas) %in%
                          as.character(covariates_geno_ctpc_tech_cc_y1_sub$sampleid)]

#check size
dim(y1_betas_cc) #424644    92

#remove probes that were used for PBMC cell type predict from age one betas
y1_betas_cc_sub <- 
  y1_betas_cc[!rownames(y1_betas_cc) %in%
                rownames(child_fscbc_ecc2_cb_probelist_deconvo), ]

#check size
dim(y1_betas_cc_sub) # 424079    92
# 565 probes removed

#check that covariates and betas in same order
identical(colnames(y1_betas_cc_sub), 
          as.character(covariates_geno_ctpc_tech_cc_y1_sub$sampleid)) # TRUE
```


Create model matrix for age one analysis
```{r model_matrix}
# create model matrix
y1_mod <- model.matrix(~ no2_preg_entire + sex + pss_6mos + 
                         maternal_smoking_y1 + maternal_education_length +
                         pbmc.PC1 + pbmc.PC2 + pbmc.PC3   +
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3,
                       data=covariates_geno_ctpc_tech_cc_y1_sub)

# check size
dim(y1_mod) # 92 13
```


Calculate the number of SVs needed for age one model.
```{r y1_sv_analysis}

# create null matrix
# this matrix does not include any of our covariates 
# this means that our covariates are "protected" and SVs will not capture them
y1_mod0 <- model.matrix(~  sex + pss_6mos + 
                          maternal_smoking_y1 + maternal_education_length +
                          pbmc.PC1 + pbmc.PC2 + pbmc.PC3 +
                          genotyping.PC1 + genotyping.PC2 + genotyping.PC3, 
                        data=covariates_geno_ctpc_tech_cc_y1_sub)

################
# sva analysis #
################

##################################
# numsv function internal to sva #
##################################

#determine the number of required SVs
#this is numsv function code pulled out

set.seed(1234)

#this is what is fed into this function from the sva function
dat = y1_betas_cc_sub
mod = y1_mod
mod0 = y1_mod0
#method="be" # this is the default and what is called by numsv
#vfilter=NULL # we dont apply a filter and its not used for be method
B=20


# initialize some stuff
# get the residual matrix (the remaining variance)
n <- ncol(dat)
m <- nrow(dat)
H <- mod %*% solve(t(mod) %*% mod) %*% t(mod)
res <- dat - t(H %*% t(dat))

# run svd and get some eigenvalues
uu <- svd(res)
# get the num of "sig" eigenvalues
ndf <- min(m, n) - ceiling(sum(diag(H)))
# get percent variance explained by each eigenvalue
dstat <- uu$d[1:ndf]^2/sum(uu$d[1:ndf]^2)
# initialize null distribution matrix
dstat0 <- matrix(0, nrow = B, ncol = ndf)

# set seed here since we are calling sample
# create the null distribution
for (i in 1:B) {
  # this part of the code removes all relationships in the 
  res0 <- t(apply(res, 1, sample, replace = FALSE))
  res0 <- res0 - t(H %*% t(res0))
  # calculate svs on this
  uu0 <- svd(res0)
  # bind their explained variance in null matrix
  dstat0[i, ] <- uu0$d[1:ndf]^2/sum(uu0$d[1:ndf]^2)
  print(i) # print out the iteration we are on
}

# determine the number of svs required here based on pvalues
psv <- rep(1, n)
for (i in 1:ndf) {
  psv[i] <- mean(dstat0[, i] >= dstat[i])
}
for (i in 2:ndf) {
  psv[i] <- max(psv[(i - 1)], psv[i])
}

# this is the variable holding the number of SVs required 
# number is each to all the svs with a pval less than/equal to 0.1
nsv <- sum(psv <= 0.1)


####################################
# code from the main SVA function  #
####################################

n.sv = nsv # from code above (run internally when sva called)
B = 5 # number of iterations

# dat =  betas # already initialized
# mod = mod  # already initialized
# mod0 = mod0  # already initialized 
# controls = NULL # we dont include controls - only for supervised method
# method = "irw" # this isnt actually used but would be called by sva
# vfilter = NULL # we dont set this when running sva
# numSVmethod = "be" # this is the default called by sva (code above)


###################################
# irwsva function internal to sva #
###################################

# this code is really want does all the work in sva
# we use the irw method so this is the code for that method
# there are also supervised and two step methods (not used by us)

# irw function takes input from sva, so we have already defined below vars
# dat = dat 
# mod = mod
# mod0 = mod0
# n.sv = n.sv
# B = B

# intialize variables again
n <- ncol(dat)
m <- nrow(dat)
Id <- diag(n)
resid <- dat %*% (Id - mod %*% solve(t(mod) %*% mod) %*% t(mod))
uu <- eigen(t(resid) %*% resid)
vv <- uu$vectors
ndf <- n - dim(mod)[2]
pprob <- rep(1, m)
one <- rep(1, n)
Id <- diag(n)
df1 <- dim(mod)[2] + n.sv
df0 <- dim(mod0)[2] + n.sv
rm(resid)

cat(paste("Iteration (out of", B, "):"))
for (i in 1:B) {
  mod.b <- cbind(mod, uu$vectors[, 1:n.sv])
  mod0.b <- cbind(mod0, uu$vectors[, 1:n.sv])
  ptmp <- f.pvalue(dat, mod.b, mod0.b)
  pprob.b <- (1 - edge.lfdr(ptmp))
  mod.gam <- cbind(mod0, uu$vectors[, 1:n.sv])
  mod0.gam <- cbind(mod0)
  ptmp <- f.pvalue(dat, mod.gam, mod0.gam)
  pprob.gam <- (1 - edge.lfdr(ptmp))
  pprob <- pprob.gam * (1 - pprob.b)
  dats <- dat * pprob
  dats <- dats - rowMeans(dats)
  uu <- eigen(t(dats) %*% dats)
  cat(paste(i, " "))
}

# this is what is normally returned by sva
# problem is that it doesnt include the diagonal matrix we want
sv_y1 = svd(dats)$v[, 1:n.sv, drop = FALSE] 
svobj_y1 <- list(sv = sv_y1, pprob.gam = pprob.gam, pprob.b = pprob.b, 
                 n.sv = n.sv)

# to get all info (including diagonal)
svobj_all_y1 <- svd(dats)

# populate rownames with sample ids
colnames(svobj_all_y1$v) <- covariates_geno_ctpc_tech_cc_y1_sub$sampleid


# save 
save(sv_y1, file = here("output_data", "linear_regression", "2022-05-06_ageone_svs.RData"))

###############
#scree of SVs #
###############

# from elodie portales-casamar

# heat scree prepartion
vars <- svobj_all_y1[['d']]^2
vars <- vars/sum(vars)
names(vars) <- paste0('SV', 1:length(vars))
xp <- 1:svobj$n.sv
varsdf <- as.data.frame(vars)
varsdf$svnum <- rep(1:length(vars))

# heat scree ggplot!! 
y1_sv_scree <- ggplot(varsdf[1:20,], aes(x=svnum, y=vars*100)) +
  geom_col() +
  theme_bw() +
  labs(x="Surrogate Variables", y = "Percentage of Variance")

y1_sv_scree

ggsave(plot = y1_sv_scree, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal", 
                       "2022-05-06_y1_sv_heat_scree.png"),
       height = 5, width = 15, units = "cm")


######################
# correlation matrix #
######################

# add svs to covariates
covariates_geno_ctpc_tech_cc_y1_sub <- cbind(covariates_geno_ctpc_tech_cc_y1_sub, sv_y1)

# rename columns with svs
colnames(covariates_geno_ctpc_tech_cc_y1_sub)[173:186] <-
  paste("y1sv",colnames(covariates_geno_ctpc_tech_cc_y1_sub)[173:186], sep="")


# early exp and pn_18 are the same thing. 
# exclude early_exp from correlation matrices
corr_y1 <- corr.mat(covariates_geno_ctpc_tech_cc_y1_sub %>% 
                      dplyr::select(no2_preg_entire, 
                                    sex, maternal_smoking_y1,
                                    maternal_education_length, pss_6mos,
                                    pbmc.PC1, pbmc.PC2, pbmc.PC3,
                                    genotyping.PC1, genotyping.PC2,
                                    genotyping.PC3, y1_chip, y1_run, y1_row, 
                                    y1sv1, y1sv2, y1sv3, y1sv4, y1sv5, y1sv6, 
                                    y1sv7, y1sv8, y1sv9, y1sv10, y1sv11, 
                                    y1sv12, y1sv13, y1sv14) %>%
                      rename("Average prenatal NO2" = no2_preg_entire,
                             Sex = sex,
                             "Year one maternal smoking" = maternal_smoking_y1,
                             "Maternal education (years)" = maternal_education_length,
                             "Maternal perceived stress 6 months" = pss_6mos,
                             "PBMC PC1" = pbmc.PC1, "PBMC PC2" = pbmc.PC2,
                             "PBMC PC3" = pbmc.PC3, 
                             Chip = y1_chip, Run = y1_run, Row = y1_row,
                             SV1 = y1sv1, SV2 = y1sv2, SV3 = y1sv3,
                             SV4 = y1sv4, SV5 = y1sv5, SV6 = y1sv6,
                             SV7 = y1sv7, SV8 = y1sv8, SV9 = y1sv9,
                             SV10 = y1sv10, SV11 = y1sv11, SV12 = y1sv12,
                             SV13 = y1sv13, SV14 = y1sv14))

pval_y1 <- pval.mat(covariates_geno_ctpc_tech_cc_y1_sub %>% 
                      dplyr::select(no2_preg_entire, 
                                    sex, maternal_smoking_y1,
                                    maternal_education_length, pss_6mos,
                                    pbmc.PC1, pbmc.PC2, pbmc.PC3,
                                    genotyping.PC1, genotyping.PC2,
                                    genotyping.PC3, y1_chip, y1_run, y1_row, 
                                    y1sv1, y1sv2, y1sv3, y1sv4, y1sv5, y1sv6, 
                                    y1sv7, y1sv8, y1sv9, y1sv10, y1sv11, 
                                    y1sv12, y1sv13, y1sv14) %>%
                      rename("Average prenatal NO2" = no2_preg_entire,
                             Sex = sex,
                             "Year one maternal smoking" = maternal_smoking_y1,
                             "Maternal education (years)" = maternal_education_length,
                             "Maternal perceived stress 6 months" = pss_6mos,
                             "PBMC PC1" = pbmc.PC1, "PBMC PC2" = pbmc.PC2,
                             "PBMC PC3" = pbmc.PC3, 
                             Chip = y1_chip, Run = y1_run, Row = y1_row,
                             SV1 = y1sv1, SV2 = y1sv2, SV3 = y1sv3,
                             SV4 = y1sv4, SV5 = y1sv5, SV6 = y1sv6,
                             SV7 = y1sv7, SV8 = y1sv8, SV9 = y1sv9,
                             SV10 = y1sv10, SV11 = y1sv11, SV12 = y1sv12,
                             SV13 = y1sv13, SV14 = y1sv14))

svcorr_y1 <- corr.plot(pval = pval_y1, 
                       corr = corr_y1, 
                       label.sig = T,
                       shape = 8,shape.size = 1,
                       axis.text.size = 8, x.axis.text.angle = 90) +
  scale_y_discrete(position = "right") +
  theme(legend.position = "none",
        panel.grid = element_blank())
svcorr_y1

ggsave(plot = svcorr_y1,
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-06_y1_sv_correlations.png"),
       height = 15,
       width = 15, 
       units = "cm")


######################################
# heat scree plot of svs against pcs #
######################################

#convert raw betas to mvals
mvals <- lumi::beta2m(y1_betas_cc_sub)
max(mvals)
min(mvals)

#pca 
uncor.dat <- t(scale(t(mvals)))
PCA_full<-princomp(na.omit(uncor.dat))
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance <-vars/sum(vars)

#Specify which covariates are categorical and/or continuous
meta_categorical <- 
  covariates_geno_ctpc_tech_cc_y1_sub[,c("sex", "maternal_smoking_y1",
                                         "y1_run", "y1_chip")] 
meta_continuous <- 
  covariates_geno_ctpc_tech_cc_y1_sub[,c("no2_preg_entire",
                                         "maternal_education_length", 
                                         "pss_6mos", "csed_6mos","y1_row",
                                         "pbmc.PC1", "pbmc.PC2", "pbmc.PC3",
                                         "pbmc.PC4",
                                         "genotyping.PC1", "genotyping.PC2",
                                         "genotyping.PC3", "sv1", "sv2", "sv3",
                                         "sv4", "sv5", "sv6", "sv7", "sv8",
                                         "sv9", "sv10", "sv11", "sv12", "sv13",
                                         "sv14", "y1_row")]

# Specify the number of PCs you want shown 
# pick enough to view most of the variance
Num<-10

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-c(1:40)

#Apply function on PCA results of uncorrected data, pulls in the meta data and beta values from above
heat_scree_plot(Loadings = Loadings, 
                Importance = Importance, 
                Num = Num, 
                Order = Order,
                Categorical = meta_categorical,
                Continuous = meta_continuous)
```



Examine persistence of prenatal changes at age one.
```{r persistence}

##############
# cg27481559 #
##############

# add dnam to covariates for linear regression
covariates_geno_ctpc_tech_cc_y1_sub$y1_cg27481559 <- 
  y1_betas_cc_sub[rownames(y1_betas_cc_sub)=="cg27481559", ]

# linear regression
y1_cg27481559_lm <- lm(y1_cg27481559 ~  
                         no2_preg_entire + sex + pss_6mos +
                         maternal_smoking_y1 + maternal_education_length +
                         pbmc.PC1 + pbmc.PC2 + pbmc.PC3 +
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 +
                         y1sv1 + y1sv2 + y1sv3 + y1sv4 + y1sv5 +  y1sv6 + 
                         y1sv7 + y1sv8 + y1sv9 + y1sv10 + y1sv11 + y1sv12 + 
                         y1sv13 + y1sv14, 
                       data=covariates_geno_ctpc_tech_cc_y1_sub)

# summary of lm
summary(y1_cg27481559_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
y1_cg27481559_no2_me <- ggpredict(y1_cg27481559_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("y1_cg27481559" = predicted,
         "no2_preg_entire" = x)

# plot 
y1_cg27481559_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc_y1_sub,
                                  aes(x = no2_preg_entire, 
                                      y= y1_cg27481559)) + 
  geom_line(data = y1_cg27481559_no2_me, 
            aes(x=no2_preg_entire, y = y1_cg27481559),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = y1_cg27481559_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  scale_y_continuous(limits = c(0.15,0.45), 
                     breaks=seq(0.20,0.40,0.10),
                     labels = scales::number_format(accuracy = 0.01)) +
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = y1_cg27481559_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_y1_cg27481559_dnam.png"),
       width = 3.7, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc_y1_sub$y1_cg27481559, 
    covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire,
    method="pearson") # -0.1177425

# effect size
range_y1 = max(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire)

summary(y1_cg27481559_lm)$coefficients["no2_preg_entire", "Estimate"]*range_y1 # -0.1103026


#############################
# AFAP1/LOC84740 cg12228123 #
#############################

#add dnam to covariates for linear regression
covariates_geno_ctpc_tech_cc_y1_sub$y1_cg12228123<- 
  y1_betas_cc_sub[rownames(y1_betas_cc_sub)=="cg12228123", ]

#linear regression
y1_cg12228123_lm <- lm(y1_cg12228123 ~ 
                         no2_preg_entire + sex + pss_6mos +
                         maternal_smoking_y1 + maternal_education_length +
                         pbmc.PC1 + pbmc.PC2 + pbmc.PC3 +
                         genotyping.PC1 + genotyping.PC2 + genotyping.PC3 +
                         y1sv1 + y1sv2 + y1sv3 + y1sv4 + y1sv5 +  y1sv6 + 
                         y1sv7 + y1sv8 + y1sv9 + y1sv10 + y1sv11 + y1sv12 + 
                         y1sv13 + y1sv14, 
                       data=covariates_geno_ctpc_tech_cc_y1_sub)

# summary of lm
summary(y1_cg12228123_lm)

# estimate marginal effects of prenatal NO2 exposure on DNAm
y1_cg12228123_no2_me <- ggpredict(y1_cg12228123_lm, terms = "no2_preg_entire") %>%
  as.data.frame() %>% 
  rename("y1_cg12228123" = predicted,
         "no2_preg_entire" = x)

# plot 
y1_cg12228123_dnam_plot <- ggplot(covariates_geno_ctpc_tech_cc_y1_sub,
                                  aes(x = no2_preg_entire, 
                                      y= y1_cg12228123)) + 
  geom_line(data = y1_cg12228123_no2_me, 
            aes(x=no2_preg_entire, y = y1_cg12228123),
            col="red", size=0.5, linetype="solid") +
  geom_ribbon(data = y1_cg12228123_no2_me, 
              aes(x = no2_preg_entire, 
                  ymin = conf.low, ymax = conf.high), 
              alpha = .2) + 
  scale_y_continuous(limits = c(0.50,1.0), breaks=seq(0.55,0.95,0.20)) +
  geom_point(alpha=0.6, col = "#254A90", size=1) + 
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=9),
        panel.border = element_blank(),
        axis.line = element_line(color ="black"),
        axis.title = element_blank()) 

ggsave(plot = y1_cg12228123_dnam_plot, 
       filename = here("figures", 
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_y1_cg12228123_dnam.png"),
       width = 3.7, height = 3.2, units = "cm")

# correlation 
cor(covariates_geno_ctpc_tech_cc_y1_sub$y1_cg12228123, 
    covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire,
    method="pearson") # -0.228647

# effect size
range_y1 = max(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire)

summary(y1_cg12228123_lm)$coefficients["no2_preg_entire", "Estimate"]*range_y1 # -0.1103026
```

Examine how effect size changes over first year of life at top 100 cord blood cpgs.
```{r top100_effect_size_persistence}

# get top 100 cord blood cpgs 
cb100 <- cb_pvals_tech %>% 
  subset(db > 0.025 |  db < -0.025) %>%
  arrange(pval_no2_preg_entire_bacon_adj) %>%
  slice_head(n=100)

rownames(cb100) <- cb100$cpgid

# same results as above (just double checking)
# get top 100 cord blood cpgs 
# cb1001 <- cb_pvals_tech %>% 
#     subset(db > 0.025 |  db < -0.025) %>%
#     arrange(pval_no2_preg_entire_bacon) %>%
#     slice_head(n=100)


# get age one dnam for top 100 cord blood cpgs
pb100_betas <- y1_betas_cc_sub[rownames(cb100),]

# check size
dim(pb100_betas)

# check order
identical(rownames(cb100), rownames(pb100_betas)) # TRUE

# transpose for db calculations
pb100_betas_t <- t(pb100_betas)

# calculate delta beta for each of the 100 age one cpgs
pb100_pvals <- 
  pbapply(pb100_betas_t, 2, function(x) {
    
    combined <- cbind(covariates_geno_ctpc_tech_cc_y1_sub, x)
    
    #run the linear regression on this cpg using all 5 imputed covariate data sets
    fit <- lm(x ~ no2_preg_entire + sex + pss_6mos +
                maternal_smoking_y1 + maternal_education_length +
                pbmc.PC1 + pbmc.PC2 + pbmc.PC3 +
                genotyping.PC1 + genotyping.PC2 + genotyping.PC3 +
                y1sv1 + y1sv2 + y1sv3 + y1sv4 + y1sv5 +  y1sv6 + 
                y1sv7 + y1sv8 + y1sv9 + y1sv10 + y1sv11 + y1sv12 + 
                y1sv13 + y1sv14, data=combined)
    
    #get summary of lm fit
    #this includes coefficients and pvalues
    #coerce into df for manipulation
    #add in terms (rownames) as column for annotation
    fit_sum <- summary(fit)
    fit_sum_coeff <- as.data.frame(fit_sum$coefficients)
    fit_sum_coeff$term <- rownames(fit_sum_coeff)
    
    #this data frame has rows for each linear regression variable, and columns for statistics
    #want a dataframe of 1 row and many columns (one for each variable with each statistic type)
    #do this by using the spread function for each statistic
    #add cpg id onto dataframe for pvalues (to keep track of information)
    #then cbind dataframes together
    
    #coefficient
    estimate <- spread(fit_sum_coeff[,c("term", "Estimate")], term, Estimate)
    colnames(estimate) <- paste("estimate", colnames(estimate), sep=".")
    
    #std error
    std.error <- spread(fit_sum_coeff[,c("term", "Std. Error")], term, 'Std. Error')
    colnames(std.error) <- paste("std.error", colnames(std.error), sep=".")
    
    #t statistics
    statistic <- spread(fit_sum_coeff[,c("term", "t value")], term, 't value')
    colnames(statistic) <- paste("t.val", colnames(statistic), sep=".")
    
    #pvalue
    p.value <- spread(fit_sum_coeff[,c("term", "Pr(>|t|)")], term, 'Pr(>|t|)')
    colnames(p.value) <- paste("p.value", colnames(p.value), sep=".")
    
    #add in cpgid for annotation 
    p.value$cpg <- colnames(x)
    
    #cbind into wide dataframe
    widepool <- cbind(estimate, std.error, statistic, p.value)
    
    #return the wide dataframe
    #these will be do.call(rbind()) together after
    return(widepool)
    
  })

# rbind list to data frame 
pb100_pvals_df <- do.call(rbind, pb100_pvals)

# calculate effect size at age one
range_y1 = max(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire) -
  min(covariates_geno_ctpc_tech_cc_y1_sub$no2_preg_entire)

pb100_pvals_df$db <- pb100_pvals_df$estimate.no2_preg_entire * range_y1

pb100_pvals_df$p.value.no2_preg_entire_adj <- p.adjust(pb100_pvals_df$p.value.no2_preg_entire, method = "bonferroni")

# check that rownames match between cord blood and peripheral blood
identical(rownames(cb100), rownames(pb100_pvals_df)) # TRUE

# bind age one effect size to cord blood data frame
cb100$db_y1 <- pb100_pvals_df$db
cb100$pval_no2_preg_entire_y1 <- pb100_pvals_df$p.value.no2_preg_entire*2

# calculate the change in db between birth and age one
cb100$delta_db <- cb100$db_y1 - cb100$db

# change to long format for plotting 
cb100_long <- cb100 %>% 
  mutate(dnam_dir = ifelse(db >= 0, 
                           "Higher DNAm at birth", 
                           "Lower DNAm at birth"),
         db_cb = db) %>%
  pivot_longer(cols = c(db, db_y1)) %>%
  mutate(point_sig = ifelse(pval_no2_preg_entire_bacon_adj < 0.25 & name == "db" & value < -0.025 |
                              pval_no2_preg_entire_bacon_adj < 0.25 & name == "db" & value > 0.025 |
                              pval_no2_preg_entire_bacon_adj < 0.25 & pval_no2_preg_entire_y1 < 0.05 &
                              name == "db_y1" & db_cb < -0.025 & value < -0.025 |
                              pval_no2_preg_entire_bacon_adj < 0.25 & pval_no2_preg_entire_y1 < 0.05 &
                              name == "db_y1" & db_cb > 0.025 & value > 0.025,
                            "red", "blue"),
         line_sig = ifelse(cpgid=="cg12228123", "red", "blue"))



# plot the cpgs exhibited lower 
top100_es_lower <- ggplot(cb100_long %>% 
                            subset(dnam_dir == "Lower DNAm at birth"), 
                          aes(x=name, y = value,
                              group = cpgid, color = line_sig)) +
  geom_point(alpha = 0.5, 
             color = cb100_long %>% 
               subset(dnam_dir == "Lower DNAm at birth") %>% 
               pull(point_sig), 
             size = 1) +
  geom_line(alpha = 0.5) +
  facet_wrap(~"Lower cord blood \nDNAm associated \nwith prenatal NO2") +
  scale_color_manual(values = c("#254A90",  "red")) +
  scale_x_discrete(labels = c("Birth", "Age one"), name = "Time point") +
  scale_y_continuous(name = "Effect size (% DNA methylation)",
                     limits = c(-0.25, 0.10),
                     breaks = seq(-0.25, 0.10, 0.05)) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        strip.background = element_blank(),
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        legend.position = "none") +
  geom_signif(comparisons = list(c("db", "db_y1")),
              colour = "black",
              y_position = 0.05,
              annotations = "p = 6.1e-16",
              vjust=-0.25,
              textsize = 3.5) 

top100_es_higher <- ggplot(cb100_long %>% 
                             subset(dnam_dir == "Higher DNAm at birth"), 
                           aes(x=name, y = value,
                               group = cpgid, color = line_sig)) +
  geom_point(alpha = 0.5, 
             color = cb100_long %>% 
               subset(dnam_dir == "Higher DNAm at birth") %>% 
               pull(point_sig),
             size = 1) +
  geom_line(alpha = 0.5) +
  facet_wrap(~"Higher cord blood \nDNAm associated \nwith prenatal NO2") +
  scale_color_manual(values = c("#254A90",  "red")) +
  scale_x_discrete(labels = c("Birth", "Age one"), name = "Time point") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(color = "black"),
        strip.background = element_blank(),
        axis.text = element_text(size=9),
        axis.title = element_blank(),
        legend.position = "none") +
  geom_signif(comparisons = list(c("db", "db_y1")),
              y_position = 0.25, 
              colour = "black",
              annotations = "p = 7.3e-12",
              vjust=-0.25,
              textsize = 3.5) +
  scale_y_continuous(limits = c(-0.05, 0.30),
                     breaks = seq(-0.05, 0.30, 0.05),
                     name= "")

top100_es <- cowplot::plot_grid(top100_es_lower, top100_es_higher, 
                                ncol=2, nrow=1)


ggsave(plot = top100_es,
       filename = here("figures",
                       "linear_regression", 
                       "prenatal",
                       "2022-05-10_top100_es_change.png"),
       height = 7, width = 8, units = "cm")


# one sample paired t-test for decreased dnam
# alternative = "greater" is the alternative that x has a larger mean than y.
t.test(x=cb100 %>% subset(db<=0) %>% pull(db_y1),
       y=cb100 %>% subset(db<=0) %>% pull(db),
       paired = TRUE, alternative = "greater")
# p-value = 6.085e-16

# one sample pairwd t-test for increased dnam
t.test(x=cb100 %>% subset(db>0) %>% pull(db_y1),
       y=cb100 %>% subset(db>0) %>% pull(db),
       paired = TRUE, alternative = "less")
# p-value = 7.275e-12
```


"Table 1" of persistence model
```{r age_one_persistence_table1}
library(table1)

# rename rownames of cbmc estimates
identical(rownames(child_fscbc_ecc2_y1_df), 
          rownames(y1_pdata)) # TRUE
child_fscbc_ecc2_y1_df$sampleid <- y1_pdata$Sample_Label

# bind cell type estimates with covariate info
covariates_geno_ctpc_tech_cc_y1_sub_pbmc <- merge(covariates_geno_ctpc_tech_cc_y1_sub, 
                                                  child_fscbc_ecc2_y1_df, 
                                                  by = "sampleid")

# table 1
table1(~sex + gestational_days + pss_6mos + no2_preg_entire + 
         maternal_education_length +  maternal_ethnicity + 
         maternal_smoking_y1 + CD4T + CD8T  + Bcell + Mono + NK,
       data = covariates_geno_ctpc_tech_cc_y1_sub_pbmc)

# prenatal no2 quartiles
quantile(covariates_geno_ctpc_tech_cc_y1_sub_pbmc$no2_preg_entire)
```




Boxplots of NO2, PM2.5
```{r airpollution_boxplots}

# prenatal no2
ggplot(covariates_geno_ctpc_tech_cc, aes(y=no2_preg_entire, x=studycentre, fill=studycentre)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height=0, width=0.25, alpha=0.8) +
  theme_bw() +
  scale_fill_manual(values=c("#254A90", "#508FC2", "#92C5EB", "white")) +
  labs(x="Study Centre", y=bquote("Prenatal" ~NO[2]~ "(ppb)")) +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size=14),
        axis.title = element_text(size=16))

# year one no2
ggplot(covariates_geno_ctpc_tech_cc_y1_sub, aes(y=no2_y1_entire, x=studycentre,fill=studycentre)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height=0, width=0.25, alpha=0.8) +
  scale_fill_manual(values=c("#254A90", "#508FC2", "#92C5EB", "white")) +
  labs(x="Study Centre", y=bquote("Year one" ~NO[2]~ "(ppb)")) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size=14),
        axis.title = element_text(size=16))

# Pm2.5 B
ggplot(covariates_geno_ctpc_tech, aes(y=PM25DALbYY_01, x=studycentre, fill=studycentre)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(height=0, width=0.25, alpha=0.8) +
  scale_fill_manual(values=c("#254A90", "#508FC2", "#92C5EB", "white")) +
  labs(x="Study Centre", y=bquote(~PM[2.5]~ "(ppb)")) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.text = element_text(size=14),
        axis.title = element_text(size=16))

```
