---
title: "CHILD comparison of deconvolution methods for CBMC and PBMCs"
author: "Samantha Lee"
date: "2021-09-24"
output: html_document
---

### Load libraries 

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(sva)
library(minfi)
library(ExperimentHub)
library(FlowSorted.Blood.EPIC)
library(FlowSorted.Blood.450k)
library(FlowSorted.CordBloodCombined.450k)
library(FlowSorted.CordBlood.450k)
library(here)
```

Cell type deconvolution allows prediction of cell type proportions in cordblood and PMBCs. Cell type proprotions are corrected for in later analysis. 

For deconvolution using estimateCellCounts2 or estimateCellCounts must use RGset (estimateCellCounts2 allows mset but functionality is reduced). CHILD rgset_nobadsamp has bad samples removed. Bad probes, cross reactive probes, xy probes, and SNP probes have not yet been removed.  

### Load DNA methylation data and separate into cord and peripheral blood

Load rgset_nobadsamp for deconvolution comparison.  
```{r load_rgset}
load(here("output_data", 
          "preprocessed_data", 
          "2022-09-27_child_rgset_nobadsamp.Rdata"))
```

Separate cordblood (CMBCs) and year 1 PMBCs.
```{r separate_tissues}
# correct sample mix up 
# sample mix up was determined in preprocessing and analysis after rgset_nobadsamp was generated
# tissue mix up for sample 20113 -> relabel
rgset_nobadsamp$SampleID[5] <- "20113_P"
rgset_nobadsamp$Tissue[5] <- "P"
rgset_nobadsamp$SampleID[6] <- "20113_C"
rgset_nobadsamp$Tissue[6] <- "C"

# cord blood 
cb_rgset <- rgset_nobadsamp[,rgset_nobadsamp$Tissue=="C"]
dim(cb_rgset) 
# 622399  153 
# this data set contains duplicates - okay as increases internal consistency 

# year 1 pmbcs
y1_rgset <- rgset_nobadsamp[,rgset_nobadsamp$Tissue=="P"]
dim(y1_rgset) 
# 622399    145
```


Load improved estimated cell counts 2 function
```{r load_estimatecellcounts2}
source(here("scripts",
            "2022-09-27_estimateCellCounts2v2.R"))
```

###  Cell type deconvolution of CBMCs (cord blood) and year 1 PBMCs using cordblood reference sets. 

Cell type correction using estimateCellCounts2 and FlowSorted.CordBloodCombined.450k - without WBCs.  
```{r child_fscbc_ecc2}
hub <- ExperimentHub()
myfiles <- query(hub, "FlowSorted.CordBloodCombined.450k")
FlowSorted.CordBloodCombined.450k <- myfiles[[1]]
FlowSorted.CordBloodCombined.450k

# CBMCs (cordblood) with cord blood reference set 
# dont include granulocytes since they are not expected in CBMCs
child_fscbc_ecc2_cb_any <- 
  estimateCellCounts2v2(rgSet = cb_rgset, 
                        compositeCellType = "CordBlood",
                        processMethod = "preprocessNoob",
                        probeSelect = "any",
                        cellTypes = c("CD4T","NK","nRBC","Bcell","Mono","CD8T"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.CordBloodCombined.450k",
                        returnAll = FALSE) 

# CBMCs (cordblood) with cord blood reference set 
# dont include granulocytes since they are not expected in CBMCs
# pick equal numbers of probes with magnitude of effect in both directions
child_fscbc_ecc2_cb_both <- 
  estimateCellCounts2v2(rgSet = cb_rgset, 
                      compositeCellType = "CordBlood",
                      processMethod = "preprocessNoob",
                      probeSelect = "both",
                      cellTypes = c("CD4T","NK","nRBC","Bcell","Mono","CD8T"), 
                      referencePlatform = "IlluminaHumanMethylation450k",
                      referenceset = "FlowSorted.CordBloodCombined.450k",
                      returnAll = FALSE) 


# year 1 PBMCs with cordblood reference set
# dont include granulocytes or nRBCs since neither are expected in year 1 PBMCs
child_fscbc_ecc2_y1_any <- 
  estimateCellCounts2v2(rgSet = y1_rgset, 
                      compositeCellType = "CordBlood",
                      processMethod = "preprocessNoob",
                      probeSelect = "any",
                      cellTypes = c("CD4T","NK", "Bcell","Mono","CD8T"), 
                      referencePlatform = "IlluminaHumanMethylation450k",
                      referenceset = "FlowSorted.CordBloodCombined.450k",
                      returnAll = FALSE) 

# year 1 PBMCs with cordblood reference set
# dont include granulocytes or nRBCs since neither are expected in year 1 PBMCs
# pick 100 probes with magnitude of effect in both direction
child_fscbc_ecc2_y1_both <- 
  estimateCellCounts2v2(rgSet = y1_rgset, 
                      compositeCellType = "CordBlood",
                      processMethod = "preprocessNoob",
                      probeSelect = "both",
                      cellTypes = c("CD4T","NK", "Bcell","Mono","CD8T"), 
                      referencePlatform = "IlluminaHumanMethylation450k",
                      referenceset = "FlowSorted.CordBloodCombined.450k",
                      returnAll = FALSE) 
```

```{r child_fscbc_ecc2_idol}
# need epic loaded to use idol cordblood set
hub <- ExperimentHub()
myfiles <- query(hub, "FlowSorted.Blood.EPIC")
FlowSorted.Blood.EPIC <- myfiles[[1]]
FlowSorted.Blood.EPIC

data("IDOLOptimizedCpGsCordBlood")

# CBMCs (cordblood) with cord blood reference set and idol probes
# dont include granulocytes since they are not expected in CBMCs
child_fscbc_ecc2_cb_idol <-
  estimateCellCounts2v2(cb_rgset, 
                        compositeCellType = "CordBlood",
                        processMethod = "preprocessNoob",
                        probeSelect = "IDOL",
                        cellTypes = c("CD4T","NK","nRBC","Bcell","Mono","CD8T"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.CordBloodCombined.450k",
                        IDOLOptimizedCpGs = IDOLOptimizedCpGsCordBlood, 
                        returnAll = FALSE) 


# year 1 PBMCs with cordblood reference set using idol probes
# # dont include granulocytes or nRBCs since neither are expected in year 1 PBMCs
child_fscbc_ecc2_y1_idol <-
  estimateCellCounts2v2(y1_rgset, 
                        compositeCellType = "CordBlood",
                        processMethod = "preprocessNoob",
                        probeSelect = "IDOL",
                        cellTypes = c("CD4T","NK", "Bcell","Mono","CD8T"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.CordBloodCombined.450k",
                        IDOLOptimizedCpGs = IDOLOptimizedCpGsCordBlood, 
                        returnAll = FALSE) 

```


###  Cell type deconvolution of year 1 PBMCs using adult reference sets. 

Cell type correction using estimateCellCounts2 and FlowSorted.Blood.EPIC.  
```{r child_fsbepic_ecc2}
# dont include neutrophils since no granulocytes are here
child_fsbepic_ecc2_y1_both <-
  estimateCellCounts2v2(y1_rgset, 
                        compositeCellType = "Blood",
                        processMethod = "preprocessNoob",
                        probeSelect = "both",
                        cellTypes = c("NK","Bcell","CD4T","CD8T","Mono"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.Blood.EPIC",
                        returnAll = FALSE) 


child_fsbepic_ecc2_y1_idol <-
  estimateCellCounts2v2(y1_rgset, 
                        compositeCellType = "Blood",
                        probeSelect = "IDOL",
                        processMethod = "preprocessNoob",
                        cellTypes = c("NK","Bcell","CD4T","CD8T","Mono"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.Blood.EPIC",
                        IDOLOptimizedCpGs = IDOLOptimizedCpGs450klegacy, 
                        returnAll = FALSE) 


child_fsbepic_ecc2_y1_any <-
  estimateCellCounts2v2(y1_rgset, 
                        compositeCellType = "Blood",
                        probeSelect = "any",
                        processMethod = "preprocessNoob",
                        cellTypes = c("NK","Bcell","CD4T","CD8T","Mono"), 
                        referencePlatform = "IlluminaHumanMethylation450k",
                        referenceset = "FlowSorted.Blood.EPIC",
                        returnAll = FALSE) 
```


Save deconvolution data.    
```{r deconvo_data}
save(child_fscbc_ecc2_cb_any,
     child_fscbc_ecc2_cb_both,
     child_fscbc_ecc2_cb_idol,
     
     child_fscbc_ecc2_y1_any,
     child_fscbc_ecc2_y1_both,
     child_fscbc_ecc2_y1_idol,
     
     child_fsbepic_ecc2_y1_any,
     child_fsbepic_ecc2_y1_both,
     child_fsbepic_ecc2_y1_idol,
     file=here("output_data", 
               "deconvolution", 
               "2022-09-28_CHILD_deconvolution_method_comparison.Rdata"))
```



