---
title: "Combined CHILD covariates, NO2, and PM2.5 data"
author: "SL"
date: "2020-11-03"
output: html_document
---

*Purpose*: The purpose of this script is to create a consolidated dataframe of all new CHILD data that was provided in the 2020 request for updated data. 

Load libraries
```{r libraries}
library(plyr)
library(tidyverse)
library(readxl)
library(here)
```

Load CHILD covariates and NO2 data
```{r load covariates}
# get list of all csv files in first set of data from 2020 request 
f <- list.files(here("input_data", 
                     "Data request Aug 2020", 
                     "CP139 data files 2"), 
                pattern = "*.csv",
                full.names = TRUE)

#read these files into a list
covariate_files = lapply(f, read.delim, sep=",")

# set names of list equal to the file names, without the file type
names(covariate_files) <- str_extract(list.files(path = here("input_data", 
                                                             "Data request Aug 2020", 
                                                             "CP139 data files 2"), 
                                                 pattern = "*.csv"), '.*(?=\\.csv)')

# read out each list item into global environment as dataframe
list2env(covariate_files, .GlobalEnv)

# get list of all csv files in second set of data from 2020 request 
g <- list.files(path = here::here("input_data", 
                                  "Data request Aug 2020", 
                                  "DR2"),
                pattern = "*.csv",
                full.names = TRUE)

# read these files into a list
covariate_files2 <- lapply(g, read.delim, sep=",")

# set names of list equal to the file names, without the file type
names(covariate_files2) <- str_extract(list.files(path = here("input_data", 
                                                              "Data request Aug 2020", 
                                                              "DR2"),
                                                  pattern = "*.csv"), '.*(?=\\.csv)')

# read out each list item into global environment as dataframe
list2env(covariate_files2, .GlobalEnv)

# old version air pollution data is in excel workbook format in second set of data from new 2020 request
h <- list.files(path = here("input_data", 
                            "Data request Aug 2020", 
                            "DR2"), 
                pattern = "NO2*",
                full.names = TRUE)

# read these files into a list
# read_xlsx usings tibbles (which suck), therefore read in as dataframe
covariate_files3 <- lapply(h, function(x) {as.data.frame(read_xlsx(x))})

# rename list items with file name
# remove file type
# replace spaces with underscore
names(covariate_files3) <- 
  sub(" ", "_", str_extract(list.files(path = here("input_data", 
                                                   "Data request Aug 2020", 
                                                   "DR2"),
                                       pattern = "NO2"), '.*(?=\\.xlsx)'))

# read out each list item into global environment as dataframe
list2env(covariate_files3, .GlobalEnv)

# read in most recent (2019 version) of no2 exposure data 
no2_2019 <- read_xlsx(here("input_data", 
                           "2022-09-26_no2_exposure_data_2019_version", 
                           "2022-09-26_NO2_exposure_pregnancy_and_first_year_of_life_2019_version.xlsx"))
no2_2019 <- as.data.frame(no2_2019)

# rename columns of no2_2019 for clarity
colnames(no2_2019)[-1] <- paste(colnames(no2_2019)[-1], "2019v", sep="_")
```

Check that all covariate variables are in same order according to CHILD REEGLE participant.
```{r check_order}
# read in order sheet
participants <- as.data.frame(read_xlsx(here("input_data", 
                                             "Data request Aug 2020", 
                                             "CP139 data files 2", 
                                             "IDs to share DR139.xlsx")))

# make list of all data frames for comparison of participant order
covariate_dfs <- Filter(function(x) is(x, "data.frame"), mget(ls()))

# return true/false for same order according to participant list
sapply(covariate_dfs, function(x) {return(identical(as.character(x[,1]), 
                                                    as.character(participants[,1])))}) # ALL TRUE

```

Combine all covariate data frames into one large master dataframe. 
```{r combine_data}
# bind all columns together
covariates <- do.call(cbind, covariate_dfs)

# remove duplicated subjectnumber columns
covariates <- dplyr::select(covariates, -contains("subjectnumber", ignore.case = TRUE))
```

Recode all missing (999) or skipped data (8888 or 888) to NA.
```{r recode_NA}
covariates[covariates == 999] <- NA
covariates[covariates == "999"] <- NA
covariates[covariates == 8888] <- NA
covariates[covariates == "8888"] <- NA
covariates[covariates == 888] <- NA
covariates[covariates == "888"] <- NA
```

Check class of data. This is important for correct covariate analysis. Refactor things with inappropriate coding.
```{r check_class}
# check that study centre is the same
studycentre <- dplyr::select(covariates, matches("studycenter", ignore.case = TRUE))

# compare if study sites are identical for each participant
identical(studycentre[,1], studycentre[,2]) #true
identical(studycentre[,1], studycentre[,3]) #true
identical(studycentre[,1], studycentre[,4]) #true
identical(studycentre[,1], studycentre[,5]) #true
identical(studycentre[,1], studycentre[,6]) #true
identical(studycentre[,1], studycentre[,7]) #true
identical(studycentre[,1], studycentre[,8]) #true
identical(studycentre[,1], studycentre[,9]) #true
identical(studycentre[,1], studycentre[,10]) #false
identical(studycentre[,1], studycentre[,11]) #true
identical(studycentre[,1], studycentre[,12]) #true
identical(studycentre[,1], studycentre[,13]) #true
identical(studycentre[,1], studycentre[,14]) #true
identical(studycentre[,1], studycentre[,15]) #false
identical(studycentre[,1], studycentre[,16]) #false
identical(studycentre[,1], studycentre[,17]) #false 
identical(studycentre[,1], studycentre[,18]) #false
identical(studycentre[,1], studycentre[,19]) #true
identical(studycentre[,1], studycentre[,20]) #true

# study centre columns which are not identical have missing values
# to be eligble for study participants had to live in same area
# can assume participants are always in the same area

# remove redundant info
covariates2 <- dplyr::select(covariates, -matches("status|study|eligibility|subjecttype", ignore.case = TRUE)) #this is info about study site and group

covariates2$studycentre <- covariates$`Q107 CBIRTHCD_cleaned.StudyCenter`

# remove these covariates as they dont add info 
covariates2 <- subset(covariates2, 
                      select= -c(urine_cotinine.cohort, # what cohort (general)
                                 urine_cotinine.visit)) # just says it was at 3 monts
# rename columns
colnames(covariates2) <- c("sex", "gestational_age", 
                           "weight_birth", "weight_3m", 
                           "weight_12m", "height_3m", 
                           "height_12m", "head_circ_3m", 
                           "head_circ_12m", "birth_month", 
                           "move_3m", "move_6m", 
                           "move_12m", "maternal_age_birth", 
                           "maternal_atopy", "prenatal_second_hand_smoke", 
                           "prenatal_maternal_smoke", "pss_pre36wk", 
                           "csed_pre36wk", "pss_pre18wk", 
                           "csed_pre18wk", "pss_6m", 
                           "csed_6m", "pss_12m", 
                           "csed_12m", "no2_prenatal_tm1_2019v", "no2_prenatal_tm2_2019v", 
                           "no2_prenatal_tm3_2019v", "no2_prenatal_avg_2019v", "no2_y1t1_2019v", 
                           "no2_y1t2_2019v", "no2_y1t3_2019v", "no2_y1t4_2019v", 
                           "no2_y1_2019v", "green_preg_2015v", "no2_prenatal_tm1_2015v", 
                           "no2_prenatal_tm2_2015v", "no2_prenatal_tm3_2015v", "no2_prenatal_avg_2015v", 
                           "no2_y1t1_2015v", "no2_y1t2_2015v", "no2_y1t3_2015v", 
                           "no2_y1t4_2015v", "no2_y1_2015v", "maternal_asthma_ever", 
                           "maternal_asthma", "paternal_asthma", 
                           "maternal_race", "paternal_race", 
                           "sample_id",  "maternal_education",
                           "maternal_education_other",
                           "maternal_education_length", 
                           "paternal_education", 
                           "paternal_education_other",
                           "paternal_education_length","total_income", 
                           "percent_income_from_gov",
                           "ses_community_rank", "ses_canada_rank",
                           "gravidity",
                           "mod", "height_birth",
                           "head_circ_birth","pets_3mos", 
                           "pests_3mos", "leaks_3mos", 
                           "mould_3mos","attached_garage_3mos",
                           "pets_6mos", "pets_y1", 
                           "pests_y1", "leaks_y1", 
                           "mould_y1", "attached_garage_y1",
                           "maternal_smoking_y1", "clinician_wheeze_dx_y1",
                           "clinician_atopy_dx_y1", "reported_asthma_dx_y1",
                           "prov_vaccines_6mos_y1", "prov_vaccines_12mos_y1",
                           "reported_wheezing_y1", "wheeze_meds_y1",
                           "prov_vaccines_2mos_6mos", "prov_vaccines_4mos_6mos",
                           "prov_vaccines_6mos_6mos", "no_vaccines_6mos",
                           "reported_asthma_dx_3y", "reported_wheeze_3y",
                           "bronchodilator_use_3y", "csi_use_3y",
                           "clinician_asthma_dx_y3", "wheeze_previous12mos_3y",
                           "clinician_wheeze_dx_y3", "clinician_allergy_dx_3y",
                           "clinican_atopy_dx_y3", "reported_asthma_dx_5y",
                           "reported_wheeze_5y", "reported_atopy_dx_5y",
                           "bronchodilator_use_5y", "csi_use_5y",
                           "clinician_asthma_dx_5y", "clinician_wheeze_wo_colds_5y",
                           "clinician_wheeze_w_colds_5y", "clinician_allergy_dx_5y",
                           "clinician_atopy_dx_5y", "maternal_reported_asthma_dx_ever_18wk",
                           "maternal_clinician_asthma_dx_ever_18wk", "maternal_smoking_18wk",
                           "pets_18wk", "pests_18wk", 
                           "mould_18wk", "leaks_18wk",
                           "attached_garage_18wk", "urine_cotinine_3m",
                           "studycentre")

# change naming for ease
covariates_old <- covariates
covariates <- covariates2

# check class
lapply(covariates, class)

# remove maternal and paternal education other class
covariates <- subset(covariates, 
                     select = -c(maternal_education_other, paternal_education_other))

# recode classes where necessary
# NB: factor variables that should be expressed in english terms (other than yes/no) are coded as character class instead of factor class

################################
#Anthropometric and ethnic data#
################################

# birth month as character
covariates$birth_month <- as.factor(covariates$birth_month)
levels(covariates$birth_month) <- c("january", "october", "november", "december",
                                    "february", "march", "april", "may",
                                    "june", "july", "august", "september")
# relevel birth to in order starting at january
covariates$birth_month <- factor(covariates$birth_month, 
                                 levels = c("january", "february", "march", "april",
                                            "may", "june", "july", "august",
                                            "september", "october", "november", "december"))

# maternal atopy factor
covariates$maternal_atopy <- as.factor(covariates$maternal_atopy)
# prenatal second hand smoke as factor
covariates$prenatal_second_hand_smoke <- as.factor(covariates$prenatal_second_hand_smoke)
# maternal smoking during prenatal period as factor
covariates$prenatal_maternal_smoke <- as.factor(covariates$prenatal_maternal_smoke)


# maternal asthma ever as factor
covariates$maternal_asthma_ever <- as.factor(covariates$maternal_asthma_ever)
# maternal asthma as factor
covariates$maternal_asthma <- as.factor(covariates$maternal_asthma)
# paternal asthma as factor
covariates$paternal_asthma <- as.factor(covariates$paternal_asthma)

# maternal race as character
covariates$maternal_race <- as.factor(covariates$maternal_race)
# paternal race as character
covariates$paternal_race <- as.factor(covariates$paternal_race)

# move 3 months
covariates$move_3m <- as.factor(covariates$move_3m)
# move 6 months
covariates$move_6m <- as.factor(covariates$move_6m)
# move 12 months
covariates$move_12m <- as.factor(covariates$move_12m)

# studycentre
covariates$studycentre <- as.factor(covariates$studycentre)

########################
#Prenatal SES (SES18WK)#
########################

# maternal education as factor
covariates$maternal_education <- as.character(covariates$maternal_education)
covariates$maternal_education <- as.factor(covariates$maternal_education)
levels(covariates$maternal_education) <- c("somehighschoool", 
                                           "completedhighschool", 
                                           "somecollege", "completedcollege",
                                           "someuniversity", 
                                           "completeduniversity",
                                           "mastersdegree", "phd")

# paternal education as factor
covariates$paternal_education <- as.character(covariates$paternal_education)
covariates$paternal_education <- as.factor(covariates$paternal_education)
levels(covariates$paternal_education) <- c("lessthanhighschool", 
                                           "somehighschoool",
                                           "completedhighschool", 
                                           "somecollege", "completedcollege",
                                           "someuniversity", 
                                           "completeduniversity",
                                           "mastersdegree", "phd")

# income as factor
covariates$total_income <- as.character(covariates$total_income)
covariates$total_income <- as.factor(covariates$total_income)
levels(covariates$total_income) <- c("10000-19999", "20000-29999", 
                                     "30000-39999", "40000-49999", 
                                     "500000-59999", "60000-79999", 
                                     "80000-99999", "100000-149999",
                                     "150000+", NA)

##############################
#Child birth chart (CBIRTHCD)#
##############################

# method of delivery as character
covariates$mod <- as.character(covariates$mod)
covariates$mod <- as.factor(covariates$mod)

# create a mod of delivery iwth just vaginal/caesarean
covariates$mod_cv <- covariates$mod

# rename levesl of mod
levels(covariates$mod) <- c("vaginal", "vaginal_forceps", "vaginal_vacuum",
                            "elective_caesarean", "caesarean_labour", 
                            "emergency_caesarean_labour",
                            "emergency_caesarean_nolabour")

# rename levels of mod_cv
levels(covariates$mod_cv) <- c("vaginal", "vaginal", "vaginal",
                               "caesarean", "caesarean", 
                               "caesarean","caesarean")


####################################
#Home environment 3 months (HENV3M)#
####################################

# furry pets at 3 months as factor
covariates$pets_3mos <- as.factor(covariates$pets_3mos)
# pests or bugs at 3 months as factor
covariates$pests_3mos <- as.factor(covariates$pests_3mos)
# water leaks at 3 months as factor
covariates$leaks_3mos <- as.factor(covariates$leaks_3mos)
# visible mould at 3 months as factor
covariates$mould_3mos <- as.factor(covariates$mould_3mos)
# attached garage at 3 months as factor
covariates$attached_garage_3mos <- as.factor(covariates$attached_garage_3mos)


####################################
#Home environment 6 months (HENV6M)#
####################################

# furry pets at 6 months as factor
covariates$pets_6mos <- as.factor(covariates$pets_6mos)

##################################
#Home environment 1 year (HENV1Y)#
##################################

# furry pets at 1 year as factor
covariates$pets_y1 <- as.factor(covariates$pets_y1)
# pests or bugs at 1 year as factor
covariates$pests_y1 <- as.factor(covariates$pests_y1)
# water leaks at 1 year as factor
covariates$leaks_y1 <- as.factor(covariates$leaks_y1)
# visible mould at 1 year as factor
covariates$mould_y1 <- as.factor(covariates$mould_y1)
# attached garage at 1 year as factor
covariates$attached_garage_y1 <- as.factor(covariates$attached_garage_y1)


##################################
#Maternal health 1 year (MHLTH1Y)#
##################################

# maternal smoking at year 1 as factor
covariates$maternal_smoking_y1 <- as.factor(covariates$maternal_smoking_y1)


################################
#Child health 1 year(CHLTH1Y)#
################################

# child asthma diagnosis at 1 year (parent reported) as factor
covariates$reported_asthma_dx_y1 <- as.factor(covariates$reported_asthma_dx_y1)
# provincial vaccines at 6 months (reported at year 1) as factor
covariates$prov_vaccines_6mos_y1 <- as.factor(covariates$prov_vaccines_6mos_y1)
# provincial vaccines at 12 months (reported at year 1) as factor
covariates$prov_vaccines_12mos_y1 <- as.factor(covariates$prov_vaccines_12mos_y1)
# reported wheeze
covariates$reported_wheezing_y1 <- as.factor(covariates$reported_wheezing_y1)
# any wheezing meds past year?
covariates$wheeze_meds_y1 <- as.factor(covariates$wheeze_meds_y1)


##############################
#Child health 1 year(CHCLA1Y)#
##############################

# clinician diagnosed wheeze
covariates$clinician_wheeze_dx_y1 <- as.factor(covariates$clinician_wheeze_dx_y1)
# clinician diagnsed atopy
covariates$clinician_atopy_dx_y1 <- as.factor(covariates$clinician_atopy_dx_y1)

############################################
#Child health questionnaire 6 mos (CHLTH6)#
############################################

# provincial vaccines at 2 months (reported at 6 months) as factor
covariates$prov_vaccines_2mos_6mos <- as.factor(covariates$prov_vaccines_2mos_6mos)
# provincial vaccines at 4 months (reported at 6 months) as factor
covariates$prov_vaccines_4mos_6mos <- as.factor(covariates$prov_vaccines_2mos_6mos)
# provincial vaccines at 6 months (reported at 6 months) as factor
covariates$prov_vaccines_6mos_6mos <- as.factor(covariates$prov_vaccines_6mos_6mos)
# no provincial vaccines (reported at 6 months) as factor
covariates$no_vaccines_6mos <- as.factor(covariates$no_vaccines_6mos)


##############################################
#Child health questionnaire 3 years (CHLTH3Y)#
##############################################

# child wheezing with or without cold as factor
covariates$reported_wheeze_3y <- as.factor(covariates$reported_wheeze_3y)
# child ever diagnosed with asthma at 3 years (maternal reported) as factor
covariates$reported_asthma_dx_3y <- as.factor(covariates$reported_asthma_dx_3y)
# bronchodilator use in last 6 months
covariates$bronchodilator_use_3y <- as.factor(covariates$bronchodilator_use_3y)
# steroid inhaler use in last 6 months
covariates$csi_use_3y <- as.factor(covariates$csi_use_3y)


####################################################
#Child clinical health assessment 3 years (CHCLA3Y)#
####################################################

# physician diagnosed asthma at 3 years as factor
covariates$clinician_asthma_dx_y3 <- as.factor(covariates$clinician_asthma_dx_y3)
# wheeze reported in last 12 months
covariates$wheeze_previous12mos_3y <- as.factor(covariates$wheeze_previous12mos_3y)
#clinician reported wheeze
covariates$clinician_wheeze_dx_y3 <- as.factor(covariates$clinician_wheeze_dx_y3)
# clinician reported atopy
covariates$clinican_atopy_dx_y3 <- as.factor(covariates$clinican_atopy_dx_y3)
# clinician reported allergy
covariates$clinician_allergy_dx_3y <- as.factor(covariates$clinician_allergy_dx_3y)


##############################################
#Child health questionnaire 5 years (CHLTH5Y)#
##############################################

# child ever diagnosed with asthma at 5 years (maternal reported) as factor
covariates$reported_asthma_dx_5y <- as.factor(covariates$reported_asthma_dx_5y)
# reported atopy
covariates$reported_atopy_dx_5y <- as.factor(covariates$reported_atopy_dx_5y)
# wheeze withor without cold
covariates$reported_wheeze_5y <- as.factor(covariates$reported_wheeze_5y)
# bronchodilator use over last year
covariates$bronchodilator_use_5y <- as.factor(covariates$bronchodilator_use_5y)
# corticosteroid inhaler use over last year
covariates$csi_use_5y <- as.factor(covariates$csi_use_5y)


####################################################
#Child clinical health assessment 5 years (CHCLA5Y)#
####################################################

# physician diagnsed asthma at 5 years as factor
covariates$clinician_asthma_dx_5y <- as.factor(covariates$clinician_asthma_dx_5y)
# clinician diagnosed stopy 
covariates$clinician_atopy_dx_5y <- as.factor(covariates$clinician_atopy_dx_5y)
# clinician diagnosed allergy
covariates$clinician_allergy_dx_5y <- as.factor(covariates$clinician_allergy_dx_5y)
# wheeze without cold
covariates$clinician_wheeze_wo_colds_5y <- as.factor(covariates$clinician_wheeze_wo_colds_5y)
# wheeze with colds
covariates$clinician_wheeze_w_colds_5y <- as.factor(covariates$clinician_wheeze_w_colds_5y)


###############################################
#Prenatal maternal health at 18 wk (PRNMH18WK)#
###############################################

# maternal asthma ever
covariates$maternal_reported_asthma_dx_ever_18wk <- as.factor(covariates$maternal_reported_asthma_dx_ever_18wk)
# maternal asthma diagnosis by doctor
covariates$maternal_clinician_asthma_dx_ever_18wk <- as.factor(covariates$maternal_clinician_asthma_dx_ever_18wk)
# current maternal smoking
covariates$maternal_smoking_18wk <- as.factor(covariates$maternal_smoking_18wk)


######################################
#Prenatal home environment (HENV18WK)#
######################################

# furry pets in home at 18wks as factor
covariates$pets_18wk <- as.factor(covariates$pets_18wk)
# pests or bugs in home at 18wks as factor
covariates$pests_18wk <- as.factor(covariates$pests_18wk)
# visible mould in home at 18wks as factor
covariates$mould_18wk <- as.factor(covariates$mould_18wk)
# water leaks at 18wks as factor
covariates$leaks_18wk <- as.factor(covariates$leaks_18wk)
# attacked garage at 18wks as factor
covariates$attached_garage_18wk <- as.factor(covariates$attached_garage_18wk)
```


Create composite/derived variables for SES, depression, stress, home environment variables(pets, pests, leaks, mould, attached garage), maternal smoking ever (prenatal period + year 1).
```{r create_composite_vars}
#######################################
#maternal stress derived variables#
#######################################

# average of pss z-scores during prenatal period (18 wk and 36 wk)
pss_18wk_z <- scale(covariates$pss_pre18wk, center = T, scale = T)
pss_36wk_z <- scale(covariates$pss_pre36wk, center = T, scale = T)

covariates$pss_prenatal <- rowMeans(data.frame(pss_18wk_z, pss_36wk_z))

# average of pss z-scores during first year of life (6 mos and 12 mos)
pss_6mos_z <- scale(covariates$pss_6m, center = T, scale = T)
pss_12mos_z <- scale(covariates$pss_12m, center = T, scale = T)

covariates$pss_y1 <- rowMeans(data.frame(pss_6mos_z, pss_12mos_z))


#######################################
#maternal depression derived variables#
#######################################

# average of csed z-scores during prenatal period (18 wk and 36 wk)
csed_18wk_z <- scale(covariates$csed_pre18wk, center = T, scale = T)
csed_36wk_z <- scale(covariates$csed_pre36wk, center = T, scale = T)

covariates$csed_prenatal <- rowMeans(data.frame(csed_18wk_z, csed_36wk_z))

# average of csed z-scores during first year of life (6 mos and 12 mos)
csed_6mos_z <- scale(covariates$csed_6m, center = T, scale = T)
csed_12mos_z <- scale(covariates$csed_12m, center = T, scale = T)

covariates$csed_y1 <- rowMeans(data.frame(csed_6mos_z, csed_12mos_z))

########################
#SES composite variable#
########################

# average of ses variables (measured at 18 wk visit) 
maternal_education_length_z <- scale(covariates$maternal_education_length, 
                                     center = T, 
                                     scale = T)

paternal_education_length_z <- scale(covariates$paternal_education_length, 
                                     center = T, 
                                     scale = T)

total_income_z <- scale(as.numeric(covariates$total_income), 
                        center = T, 
                        scale = T)

ses_community_rank_z <- scale(covariates$ses_community_rank, 
                              center = T, 
                              scale = T)

ses_canada_rank_z <- scale(covariates$ses_canada_rank, 
                           center = T, 
                           scale = T)

covariates$composite_ses <- rowMeans(data.frame(maternal_education_length_z,
                                                paternal_education_length_z,
                                                total_income_z,
                                                ses_community_rank_z,
                                                ses_canada_rank_z))

```

Save covariate matrix to rdata for future use.
```{r save_covariates}
save(covariates, 
     file=here("output_data", 
               "covariates", 
               "2022-09-27_CHILD_all_covariates.Rdata"))
```