---
title: "Age one blood cell cell type PC generation"
author: "SL"
date: "2021-09-24"
output: html_document
---

*Purpose*: This script generates cell type PCs from PBMCs. PBMC PCs can be used in linear models to correct for cell type.

## Cell type correction

Correct cordblood for cell type using linear regression. 

Based on "Adjusting for cell type composition in DNA methylation data using a regression-based approach" by Jones et al (2017).  

The data is composite data and cell types are not independent of one another:  

Two ways of dealing with this:  

1. Do a PCA  
* variables are not independent -PCA will give all the same data but in independent vectors  
* But cant do PCA on composite data - this is composite data  
* Do a centred log ratio before PCA to get around this  

2. Leave out one cell type  
* More simple way  
* Usually leave cell type that is largest proportion  
* For whole blood/cord blood - drop granulocytes  
* For PMBCs and y1MCs - drop CD4T cells  


## Principle Component (PC) generation

Load required libraries.
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(minfi)
library(rgr) #for log centred ratio transfromation
library(factoextra)
library(GGally)
library(here)
```

Load CHILD145 deconvolution data. This deconvolution does not include granulocytes as granulocytes are not expected.
```{r load_CHILD_data, message=FALSE, warning=FALSE}
load(here("output_data", 
          "deconvolution", 
          "2022-09-28_cbmc_pbmc_deconvolution.Rdata"))

#only want to keep y1 data
rm(child_fscbc_ecc2_cb)

#convert deconvolution data to dataframe
child_fscbc_ecc2_y1_df <- as.data.frame(child_fscbc_ecc2_y1$prop)
```

Use ILR and robust PCA as there are outliers in nRBCs
```{r ilr_robustPCA}
y1_ilr <- suppressWarnings(princomp(child_fscbc_ecc2_y1_df, 
                                    covmat = MASS::cov.rob(child_fscbc_ecc2_y1_df), 
                                    cor = FALSE))
```

Examine variance captured by PCs.
```{r pc_variance}
# scree plot of variance
var_explained = y1_ilr$sdev^2 /sum(y1_ilr$sdev^2)
ggplot(var_explained %>% as.data.frame(), aes(y=., x=(c(1:5)))) + geom_col()
cumsum(var_explained) # use 3 cell type PCs
#   Comp.1    Comp.2    Comp.3    Comp.4    Comp.5 
# 0.5342202 0.8068605 0.9608931 0.9986697 1.0000000 
```

Save y1MC PCs. 
```{r save_adjusted_betas}
save(y1_ilr,
     file=here("output_data", 
               "deconvolution", 
               "2022-09-28_PBMC_PCs.RData"))
```
 







