---
title: "CHILD visual comparison of deconvolution methods for CBMCs and PBMCs"
author: "SL"
date: "2020-11-02"
output: html_document
---


```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)
library(minfi)
library(reshape2)
library(ggsignif)
library(here)
```


Load deconvolution data and pData of CHILD145.
```{r load_workspace, message=FALSE, warning=FALSE}
#deconvolution data
load(here("output_data", "deconvolution", "2021-01-20_CHILD_deconvolution_method_comparison.Rdata"))

#this pdata has bad samples removed, CBMC duplicate samples removed, and sample mix-up corrected
load(here("output_data", "preprocessed_data", "2021-01-14_CHILD_preprocessed_betas_pdata_annotation.Rdata"))

#subset pData into CBMCs (cord blood) and PBMCs (year 1)
#cbmcs
cb_pdata <- subset(pdata_nodups, Tissue == "C")
#pbmcs
y1_pdata <- subset(pdata_nodups, Tissue == "P")
```


Convert CHILD cell type deconvolution data into dataframes for plotting.  
```{r convert_to_dataframes, message=FALSE, warning=FALSE}
child_fscbc_ecc2_cb_any_df <- as.data.frame(child_fscbc_ecc2_cb_any$counts)
child_fscbc_ecc2_cb_both_df <- as.data.frame(child_fscbc_ecc2_cb_both$counts)
child_fscbc_ecc2_cb_idol_df <- as.data.frame(child_fscbc_ecc2_cb_idol$counts)

child_fscbc_ecc2_y1_any_df <- as.data.frame(child_fscbc_ecc2_y1_any$counts) 
child_fscbc_ecc2_y1_both_df <- as.data.frame(child_fscbc_ecc2_y1_both$counts)
child_fscbc_ecc2_y1_idol_df <- as.data.frame(child_fscbc_ecc2_y1_idol$counts)

child_fsbepic_ecc2_y1_any_df <- as.data.frame(child_fsbepic_ecc2_y1_any$counts)
child_fsbepic_ecc2_y1_both_df <- as.data.frame(child_fsbepic_ecc2_y1_both$counts)
child_fsbepic_ecc2_y1_idol_df <- as.data.frame(child_fsbepic_ecc2_y1_idol$counts)
```


Remove duplicate samples from cord blood estimates.
```{r remove_cb_duplicates}

#make list of samples to remove from beta and pData
cb_dups <-c("9298768102_R06C01", "9341679111_R05C02", "9298768023_R03C02",
            "9341679076_R06C02", "9297962089_R06C02", "9298768023_R02C02",
            "9341679097_R02C02", "9341679114_R01C02", "9298768023_R06C01")

#remove replicates from pData                        
child_fscbc_ecc2_cb_any_df_nodups <-
  child_fscbc_ecc2_cb_any_df[!rownames(child_fscbc_ecc2_cb_any_df) %in% cb_dups, ]

child_fscbc_ecc2_cb_both_df_nodups <- 
  child_fscbc_ecc2_cb_both_df[!rownames(child_fscbc_ecc2_cb_both_df) %in% cb_dups, ]

child_fscbc_ecc2_cb_idol_df_nodups <-
  child_fscbc_ecc2_cb_idol_df[!rownames(child_fscbc_ecc2_cb_idol_df) %in% cb_dups, ]
```


## Compare cell type deconvolution of CHILD data between different methods.  

### Cord blood comparison methods


Compare cell type deconvolution of cordblood using FlowSorted.CordBloodCombined.450k (with estimateCellCounts2 and any probes) and FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and "both" probes). 
```{r child_fscbc_ecc2_cb_any_df_nodups_vs_child_fscbc_ecc2_cb_both_df_nodups}
child_fscbc_ecc2_cb_any_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_any_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_any_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.any"

child_fscbc_ecc2_cb_both_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_both_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_both_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.both"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_cb_any_df_long <- gather(child_fscbc_ecc2_cb_any_df_nodups, celltype, percentage, CD4T:CD8T)
child_fscbc_ecc2_cb_both_df_long <- gather(child_fscbc_ecc2_cb_both_df_nodups, celltype, percentage, CD4T:CD8T)


#row bind two methods together
child_fscbcany_fscbcboth <- rbind(child_fscbc_ecc2_cb_any_df_long,
                                  child_fscbc_ecc2_cb_both_df_long) 

#spread according to method
child_fscbcany_fscbcboth_wide <- spread(child_fscbcany_fscbcboth, method, percentage )


ggplot(child_fscbcany_fscbcboth_wide, 
       aes(x=FlowSorted.CordBloodCombined.450k.any,
           y=FlowSorted.CordBloodCombined.450k.both)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1, linetype="solid") +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "CordBloodCombined.450k 'both'") +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_cb_fscbcboth_fscbcany.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)

```


Compare cell type deconvolution of cordblood using FlowSorted.CordBloodCombined.450k (with estimateCellCounts2) and FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and IDOL probes).  

```{r child_fscbc_ecc2_any_df vs child_fscbc_ecc2_idol_df, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_cb_any_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_any_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_any_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.any"

child_fscbc_ecc2_cb_idol_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_idol_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_idol_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.IDOL"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_cb_any_df_long <- gather(child_fscbc_ecc2_cb_any_df_nodups, celltype, percentage, CD4T:CD8T)
child_fscbc_ecc2_cb_idol_df_long <- gather(child_fscbc_ecc2_cb_idol_df_nodups, celltype, percentage, CD4T:CD8T)

#row bind two methods together
child_fscbc_fscbcidol <- rbind(child_fscbc_ecc2_cb_any_df_long, 
                               child_fscbc_ecc2_cb_idol_df_long) 

#spread according to method
child_fscbc_fscbcidol_wide <- spread(child_fscbc_fscbcidol, method, percentage )


ggplot(child_fscbc_fscbcidol_wide, aes(x=FlowSorted.CordBloodCombined.450k.any,
                                        y=FlowSorted.CordBloodCombined.450k.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1, linetype="solid") +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "CordBloodCombined.450k 'IDOL'") +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_cb_fscbcidol_fscbcany.png",    
         path =  here("figures", "deconvolution"),
         width = 3,
         height = 2.5,
         units = "in",
         limitsize = TRUE)
```

Compare cell type deconvolution of cordblood using FlowSorted.CordBloodCombined.450k (with estimateCellCounts2 and both probes) and FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and "IDOL" probes). 
```{r child_fscbc_ecc2_cb_df_nodups_vs_child_fscbc_ecc2_cb_both_df_nodups}
child_fscbc_ecc2_cb_both_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_both_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_both_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.both"


child_fscbc_ecc2_cb_idol_df_nodups$tissue <- cb_pdata$Tissue 
child_fscbc_ecc2_cb_idol_df_nodups$sampleid <- cb_pdata$SampleID
child_fscbc_ecc2_cb_idol_df_nodups$method <- "FlowSorted.CordBloodCombined.450k.IDOL"


#convert dataframes to long format for facetting 
child_fscbc_ecc2_cb_both_df_long <- gather(child_fscbc_ecc2_cb_both_df_nodups, celltype, percentage, CD4T:CD8T)
child_fscbc_ecc2_cb_idol_df_long <- gather(child_fscbc_ecc2_cb_idol_df_nodups, celltype, percentage, CD4T:CD8T)


#row bind two methods together
child_fscbcboth_fscbcidol <- rbind(child_fscbc_ecc2_cb_both_df_long,
                                   child_fscbc_ecc2_cb_idol_df_long) 

#spread according to method
child_fscbcboth_fscbcidol_wide <- spread(child_fscbcboth_fscbcidol, method, percentage )


ggplot(child_fscbcboth_fscbcidol_wide, aes(x=FlowSorted.CordBloodCombined.450k.both,
                                        y=FlowSorted.CordBloodCombined.450k.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'both'", y = "CordBloodCombined.450k 'IDOL'") +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_cb_fscbcidol_fscbcboth.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```


Compare max and min cell type predictions in cord blood probe sets any, both, and idol.
```{r compare_max_min}
#any
print("any")
min(child_fscbc_ecc2_cb_any_df_nodups[,1:6])
max(child_fscbc_ecc2_cb_any_df_nodups[,1:6])

#both
print("both")
min(child_fscbc_ecc2_cb_both_df_nodups[,1:6])
max(child_fscbc_ecc2_cb_both_df_nodups[,1:6])

#idol
print("idol cord blood")
min(child_fscbc_ecc2_cb_idol_df_nodups[,1:6])
max(child_fscbc_ecc2_cb_idol_df_nodups[,1:6])
```


Make boxplot of cordblood nRBCs and label samples to find ones that have abnormally high predicted nRBC proportion.

```{r nRBCs_proprotions}
#from deconvolution with FlowSorted.CordBloodCombined.450k using estimateCellCounts2 (no idol probes)
ggplot(subset(child_fscbc_ecc2_cb_df_long, celltype=="nRBC"), aes(x=celltype, y=percentage)) +
  geom_boxplot(stat = "boxplot",position = "dodge2") +
  ggrepel::geom_text_repel(aes(label=sampleid))

#from deconvolution with FlowSorted.CordBloodCombined.450k using estimateCellCounts2 with IDOL probes
ggplot(subset(child_fscbc_ecc2_idol_cb_df_long, celltype=="nRBC"), aes(x=celltype, y=percentage)) +
  geom_boxplot(stat = "boxplot",position = "dodge2") +
  ggrepel::geom_text_repel(aes(label=sampleid))


#from deconvolution WITH GRANULOCYTES INCLUDED with FlowSorted.CordBloodCombined.450k using estimateCellCounts2 with IDOL probes
ggplot(subset(child_fscbc_ecc2_cb_gran_df_long, celltype=="nRBC"), aes(x=celltype, y=percentage)) +
  geom_boxplot(stat = "boxplot",position = "dodge2") +
  ggrepel::geom_text_repel(aes(label=sampleid))

```



## Year1 PBMCs comparison methods  

#### PBMCs with CordBlood Combined

Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and any probes) and FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and probeselect-borth)
```{r y1_fscbanyc_fscbcboth, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_any_df$method <- "FlowSorted.CordBloodCombined.450k.any"

#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_both_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_both_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_both_df$method <- "FlowSorted.CordBloodCombined.450k.both"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_any_df_long <- gather(child_fscbc_ecc2_y1_any_df, 
                                          celltype, percentage, CD4T:CD8T)

child_fscbc_ecc2_y1_both_df_long <- gather(child_fscbc_ecc2_y1_both_df, 
                                           celltype, percentage, CD4T:CD8T)

#row bind two methods together
child_y1_fscbcany_fscbcboth <- rbind(child_fscbc_ecc2_y1_any_df_long, child_fscbc_ecc2_y1_both_df_long) 

#spread according to method
child_y1_fscbcany_fscbcboth_wide <- spread(child_y1_fscbcany_fscbcboth, method, percentage )


ggplot(child_y1_fscbcany_fscbcboth_wide, aes(x=FlowSorted.CordBloodCombined.450k.any,
                                        y=FlowSorted.CordBloodCombined.450k.both)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "CordBloodCombined.450k 'both'") +
  scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fscbcany_fscbcboth.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```



Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and no IDOL) and FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 WITH IDOL) 

```{r y1_fscbcasny_fscbcidol, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_idol_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_idol_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_idol_df$method <- "FlowSorted.CordBloodCombined.450k.IDOL"

child_fscbc_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_any_df$method <- "FlowSorted.CordBloodCombined.450k.any"


#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_any_df_long <- gather(child_fscbc_ecc2_y1_any_df, 
                                          celltype, percentage, CD4T:CD8T)
child_fscbc_ecc2_y1_idol_df_long <- gather(child_fscbc_ecc2_y1_idol_df, 
                                           celltype, percentage, CD4T:CD8T)

#row bind two methods together
child_y1_fscbcany_fscbcidol <- rbind(child_fscbc_ecc2_y1_any_df_long, 
                                     child_fscbc_ecc2_y1_idol_df_long) 

#spread according to method
child_y1_fscbcany_fscbcidol_wide <- spread(child_y1_fscbcany_fscbcidol,
                                           method, 
                                           percentage)


ggplot(child_y1_fscbcany_fscbcidol_wide, aes(x=FlowSorted.CordBloodCombined.450k.any,
                                        y=FlowSorted.CordBloodCombined.450k.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined 'any'", y = "CordBloodCombined 'IDOL'") +
    scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fscbcany_fscbcidol.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```



Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k (using estimateCellCounts2) and FlowSorted.Blood.EPIC (using estimateCellCounts2).  

```{r y1_fscbcboth_fscbcidol, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_idol_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_idol_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_idol_df$method <- "FlowSorted.CordBloodCombined.IDOL"

child_fscbc_ecc2_y1_both_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_both_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_both_df$method <- "FlowSorted.CordBloodCombined.both"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_idol_df_long <- gather(child_fscbc_ecc2_y1_idol_df, 
                                          celltype, percentage, CD4T:CD8T)
child_fscbc_ecc2_y1_both_df_long <- gather(child_fscbc_ecc2_y1_both_df, 
                                        celltype, percentage, CD4T:CD8T)

#row bind two methods together
child_y1_fscbcidol_fscbcboth <- rbind(child_fscbc_ecc2_y1_idol_df_long,
                                child_fscbc_ecc2_y1_both_df_long) 

#spread according to method
child_y1_fscbcidol_fscbcboth_wide <- spread(child_y1_fscbcidol_fscbcboth,
                                            method, 
                                            percentage)


ggplot(child_y1_fscbcidol_fscbcboth_wide, aes(x=FlowSorted.CordBloodCombined.both,
                                        y=FlowSorted.CordBloodCombined.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'both'", y = "CordBloodCombined.450k 'IDOL'") +
    scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
  
  ggsave(filename="2021-01-20_CHILD_y1_fscbcboth_fscbidol.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```


#### PBMCs with adult blood


Compare cell type deconvolution of year1 PMBCs (no granulocytes and WITH nRBCs) between FlowSorted.Blood.EPIC (using estimateCellCounts2 with any probe) and FlowSorted.Blood.EPIC (using estimateCellCounts2 with both probes).  

```{r y1_epicany_epicboth, message=FALSE, warning=FALSE}
child_fsbepic_ecc2_y1_both_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_both_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_both_df$method <- "FlowSorted.Blood.EPIC.both"

child_fsbepic_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_any_df$method <- "FlowSorted.Blood.EPIC.any"

#convert dataframes to long format for facetting 
child_fsbepic_ecc2_y1_any_df_long <- gather(child_fsbepic_ecc2_y1_any_df, 
                                          celltype, percentage, NK:Mono)
child_fsbepic_ecc2_y1_both_df_long <- gather(child_fsbepic_ecc2_y1_both_df, 
                                             celltype, percentage, NK:Mono)

child_y1_fsbepicboth_fsbepicany <- rbind(child_fsbepic_ecc2_y1_any_df_long, child_fsbepic_ecc2_y1_both_df_long) 

#spread according to method
child_y1_fsbepicboth_fsbepicany_wide <- spread(child_y1_fsbepicboth_fsbepicany, 
                                               method, 
                                               percentage)


ggplot(child_y1_fsbepicboth_fsbepicany_wide, aes(y=FlowSorted.Blood.EPIC.both,
                                        x=FlowSorted.Blood.EPIC.any)) +
 geom_point(aes(col=celltype), size=1, alpha = 0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(y="Blood.EPIC 'both'", x = "Blood.EPIC 'any'") +
    scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-01-20_CHILD_y1_fsbepicboth_fsbepicany.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```



Compare cell type deconvolution of year1 PMBCs (no granulocytes and no nRBCs) between FlowSorted.Blood.EPIC (using estimateCellCounts2 no IDOL probes) and FlowSorted.Blood.EPIC (using estimateCellCounts2 with IDOL probes).  

```{r y1_fsbepicboth_fsbepicidol, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fsbepic_ecc2_y1_both_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_both_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_both_df$method <- "FlowSorted.Blood.EPIC.both"

child_fsbepic_ecc2_y1_idol_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_idol_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_idol_df$method <- "FlowSorted.Blood.EPIC.IDOL"

#convert dataframes to long format for facetting 
child_fsbepic_ecc2_y1_both_df_long <- gather(child_fsbepic_ecc2_y1_both_df, 
                                             celltype, percentage, NK:Mono)
child_fsbepic_ecc2_y1_idol_df_long <- gather(child_fsbepic_ecc2_y1_idol_df, 
                                             celltype, percentage, NK:Mono)

#row bind two methods together
child_y1_fsbepicboth_fsbepicidol <- rbind(child_fsbepic_ecc2_y1_both_df_long,
                                          child_fsbepic_ecc2_y1_idol_df_long) 

#spread according to method
child_y1_fsbepicboth_fsbepicidol_wide <- 
  spread(child_y1_fsbepicboth_fsbepicidol, method, percentage)


ggplot(child_y1_fsbepicboth_fsbepicidol_wide, aes(x=FlowSorted.Blood.EPIC.both,
                                        y=FlowSorted.Blood.EPIC.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="Blood.EPIC 'both'", y = "Blood.EPIC 'IDOL'") +
    scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fsbepicboth_fsbepicidol.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```




Compare cell type deconvolution of year1 PMBCs (no granulocytes and no nRBCs) between FlowSorted.Blood.EPIC (using estimateCellCounts2 with any probes) and FlowSorted.Blood.EPIC (using estimateCellCounts2 with IDOL probes).  

```{r y1_fsbepicany_fsbepicidol, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fsbepic_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_any_df$method <- "FlowSorted.Blood.EPIC.any"

child_fsbepic_ecc2_y1_idol_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_idol_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_idol_df$method <- "FlowSorted.Blood.EPIC.IDOL"

#convert dataframes to long format for facetting 
child_fsbepic_ecc2_y1_any_df_long <- gather(child_fsbepic_ecc2_y1_any_df, 
                                            celltype, percentage, NK:Mono)

child_fsbepic_ecc2_y1_idol_df_long <- gather(child_fsbepic_ecc2_y1_idol_df, 
                                             celltype, percentage, NK:Mono)

#row bind two methods together
child_y1_fsbepicany_fsbepicidol <- rbind(child_fsbepic_ecc2_y1_any_df_long,
                                         child_fsbepic_ecc2_y1_idol_df_long) 

#spread according to method
child_y1_fsbepicany_fsbepicidol_wide <- spread(child_y1_fsbepicany_fsbepicidol, 
                                               method, percentage)


ggplot(child_y1_fsbepicany_fsbepicidol_wide, aes(x=FlowSorted.Blood.EPIC.any,
                                        y=FlowSorted.Blood.EPIC.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="Blood.EPIC 'any'", y = "Blood.EPIC 'IDOL'") +
    scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fsbepicany_fsbepicidol.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```




#### PBMCs comparing CordBlood to adult blood datasets

Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k any (using estimateCellCounts2) and FlowSorted.Blood.EPIC any (using estimateCellCounts2).  
```{r y1_fscbcany_epicany, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_any_df$method <- "FlowSorted.CordBloodCombined.450k"

child_fsbepic_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_any_df$method <- "FlowSorted.Blood.EPIC.any"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_any_df_long <- gather(child_fscbc_ecc2_y1_any_df, 
                                          celltype, percentage, CD4T:CD8T)

child_fsbepic_ecc2_y1_any_df_long <- gather(child_fsbepic_ecc2_y1_any_df, 
                                        celltype, percentage, NK:Mono)

#row bind two methods together
child_y1_fscbcany_fsbepicany <- rbind(child_fscbc_ecc2_y1_any_df_long,
                                child_fsbepic_ecc2_y1_any_df_long) 

#spread according to method
child_y1_fscbcany_fsbepicany_wide <- spread(child_y1_fscbcany_fsbepicany, 
                                            method, percentage )


ggplot(child_y1_fscbcany_fsbepicany_wide, aes(x=FlowSorted.CordBloodCombined.450k,
                                        y=FlowSorted.Blood.EPIC.any)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "Blood.EPIC 'any'") +
      scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +

  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fscbcany_fsbepicany.png",
         path =  here("figures", "deconvolution"),
         width = 8,
         height = 6.5,
         units = "cm",
         limitsize = TRUE)
```


Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k any (using estimateCellCounts2) and FlowSorted.Blood.EPIC any (using estimateCellCounts2).  
```{r y1_fscbcany_epicboth, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_any_df$method <- "FlowSorted.CordBloodCombined.450k"

child_fsbepic_ecc2_y1_both_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_both_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_both_df$method <- "FlowSorted.Blood.EPIC.both"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_any_df_long <- gather(child_fscbc_ecc2_y1_any_df, 
                                          celltype, percentage, CD4T:CD8T)

child_fsbepic_ecc2_y1_both_df_long <- gather(child_fsbepic_ecc2_y1_both_df, 
                                        celltype, percentage, NK:Mono)

#row bind two methods together
child_y1_fscbcany_fsbepicboth <- rbind(child_fscbc_ecc2_y1_any_df_long,
                                child_fsbepic_ecc2_y1_both_df_long) 

#spread according to method
child_y1_fscbcany_fsbepicboth_wide <- spread(child_y1_fscbcany_fsbepicboth,
                                             method, percentage )


ggplot(child_y1_fscbcany_fsbepicboth_wide, aes(x=FlowSorted.CordBloodCombined.450k,
                                        y=FlowSorted.Blood.EPIC.both)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "Blood.EPIC 'both'") +
      scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +
  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fscbcany_fsbepicboth.png",
    path =  here("figures", "deconvolution"),
    width = 8,
    height = 6.5,
    units = "cm",
    limitsize = TRUE)
```



Compare cell type deconvolution of year1 PMBCs (no granulocytes no nRBCs) between FlowSorted.CordBloodCombined.450k (using estimateCellCounts2 and probeselect=any) and FlowSorted.Blood.EPIC (using estimateCellCounts2 with IDOL).  
```{r y1_fscbcany_fsbepicidol, message=FALSE, warning=FALSE}
#add tissue and sampleid info to child fscbc and fscb450 dataframes
child_fscbc_ecc2_y1_any_df$tissue <- y1_pdata$Tissue 
child_fscbc_ecc2_y1_any_df$sampleid <- y1_pdata$SampleID
child_fscbc_ecc2_y1_any_df$method <- "FlowSorted.CordBloodCombined.450k.any"

child_fsbepic_ecc2_y1_idol_df$tissue <- y1_pdata$Tissue 
child_fsbepic_ecc2_y1_idol_df$sampleid <- y1_pdata$SampleID
child_fsbepic_ecc2_y1_idol_df$method <- "FlowSorted.Blood.EPIC.IDOL"

#convert dataframes to long format for facetting 
child_fscbc_ecc2_y1_any_df_long <- gather(child_fscbc_ecc2_y1_any_df, 
                                          celltype, percentage, CD4T:CD8T)
child_fsbepic_ecc2_y1_idol_df_long <- gather(child_fsbepic_ecc2_y1_idol_df, 
                                             celltype, percentage, NK:Mono)

#row bind two methods together
child_y1_fscbcany_fsbepicidol <- rbind(child_fscbc_ecc2_y1_any_df_long,
                                    child_fsbepic_ecc2_y1_idol_df_long) 

#spread according to method
child_y1_fscbcany_fsbepicidol_wide <- spread(child_y1_fscbcany_fsbepicidol, 
                                             method, percentage)


ggplot(child_y1_fscbcany_fsbepicidol_wide, 
       aes(x=FlowSorted.CordBloodCombined.450k.any,
           y=FlowSorted.Blood.EPIC.IDOL)) +
  geom_point(aes(col=celltype), size=1, alpha=0.6) +
  facet_wrap( ~ celltype, ncol=2) +  
  geom_abline(intercept=0,slope=1) +
  stat_cor(method="pearson", size=2, label.x.npc = 0.35, label.y.npc = 0.1) +
  scale_color_manual(values = c("#0B5351", "#AB9B96", "#A1674A", 
                                "#BA6E6E", "#498A9E", "#931621")) + 
  labs(x="CordBloodCombined.450k 'any'", y = "Blood.EPIC 'IDOL'") +
      scale_y_continuous(breaks = seq(0, 0.65, by = 0.2), limits = c(0,0.6)) +

  theme_bw() +
  theme(legend.position ="none",
        axis.text.x = element_text(size=7),
        axis.text.y = element_text(size=7),
        axis.title.x = element_text(size=7),
        axis.title.y = element_text(size=7),
        strip.text.x = element_text(size = 6),
        panel.grid = element_blank()) +
  
  ggsave(filename="2021-04-26_CHILD_y1_fscbcany_fsbepicidol.png",
    path = here("figures", "deconvolution"),
    width = 8,
    height = 6.5,
    units = "cm",
    limitsize = TRUE)

```






Compare max and min cell type predictions in age one PBMC sets.
```{r compare_max_min_y1}
#adult ref set with probe select = both
print("FlowSorted.Blood.EPIC probeselect=both")
min(child_fsbepic_ecc2_y1_df[,1:5])
max(child_fsbepic_ecc2_y1_df[,1:5])

#adult ref set with adult idol probes
print("FlowSorted.Blood.EPIC adult IDOL probes")
min(child_fsbepic_ecc2_y1_df[,1:5])
max(child_fsbepic_ecc2_y1_df[,1:5])

#cord blood reference set with probe select = any
print("FlowSorted.CordBloodCombined.450k probeselect=any")
min(child_fscbc_ecc2_y1_df[,c(1:5)])
max(child_fscbc_ecc2_y1_df[,c(1:5)])
```





### Compare cell type proportions between age groups using boxplots

Compare cell type proportions between birth (using FlowSorted.CordBloodCombined.450k and cord blood IDOL probes) and age 1 (using FlowSorted.Blood.EPIC with adult IDOL probes).
```{r boxplot_cbidol_y1idol}
#substring sample id so that numbering is consistent between birth and age one
child_fscbc_ecc2_idol_cb_df_nodups$sampleid <- substr(child_fscbc_ecc2_idol_cb_df_nodups$sampleid, 1, 5)
child_fsbepic_ecc2_idol_y1_df$sampleid <- substr(child_fsbepic_ecc2_idol_y1_df$sampleid, 1, 5) 

#populate nRBC column in age1 group with NA
child_fsbepic_ecc2_idol_y1_df$nRBC <- NA

#rearrange columns so that they are in the same order
child_fsbepic_ecc2_idol_y1_df <- child_fsbepic_ecc2_idol_y1_df[ colnames(child_fscbc_ecc2_idol_cb_df_nodups)]
#check column order
head(child_fsbepic_ecc2_idol_y1_df)
head(child_fscbc_ecc2_idol_cb_df_nodups)

#subset age one data so that only samples with data at both birth and age one are plotted/tests
child_fsbepic_ecc2_idol_y1_df <- child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                           child_fscbc_ecc2_idol_cb_df_nodups$sampleid, ] 
#check sizes
dim(child_fsbepic_ecc2_idol_y1_df)
dim(child_fscbc_ecc2_idol_cb_df_nodups)

#check that samples are in same order
identical(child_fscbc_ecc2_idol_cb_df_nodups$sampleid, 
           child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                           child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"sampleid"]) #true

#rowbind dataframes together
cby1 <- rbind(child_fscbc_ecc2_idol_cb_df_nodups, child_fsbepic_ecc2_idol_y1_df)

#melt dataframe for plotting
cby1_long <- cby1 %>% melt(id = c("tissue", "sampleid", "method"))

#change class of variables for plotting
cby1_long$sampleid <- as.factor(cby1_long$sampleid)


#t-tests between birth and age one for cell types


#cd4t
cd4t <- t.test(child_fscbc_ecc2_idol_cb_df_nodups[,"CD4T"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"CD4T"],
               alternative = "two.sided", 
               paired=TRUE)

#cd8t
cd8t <- t.test(child_fscbc_ecc2_idol_cb_df_nodups[,"CD8T"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"CD8T"],
               alternative = "two.sided", 
               paired=TRUE)

#mono
mono <-  t.test(child_fscbc_ecc2_idol_cb_df_nodups[,"Mono"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"Mono"],
               alternative = "two.sided", 
               paired=TRUE)

#NK cells
nk <-  t.test(child_fscbc_ecc2_idol_cb_df_nodups[,"NK"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"NK"],
               alternative = "two.sided", 
               paired=TRUE)

#bcells
bcell <- t.test(child_fscbc_ecc2_idol_cb_df_nodups[,"Bcell"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_idol_cb_df_nodups$sampleid,"Bcell"],
               alternative = "two.sided", 
               paired=TRUE)


#graphing
ggplot(cby1_long, aes(x=as.factor(tissue),
                      y=value, 
                      group=sampleid, 
                      col=sampleid))  +
  geom_line() +
  geom_point(cby1_long %>% subset(variable=="nRBC"), 
             mapping=aes(x=as.factor(tissue), 
                         y=value, 
                         group=sampleid,
                         col=sampleid)) +
  facet_grid(. ~ variable) +
  theme(legend.position = "none") +
  theme_bw() +
  theme(legend.position ="none") +
  geom_signif(comparisons = list(c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P")),
              test="t.test", test.args=list(alternative="two.sided", paired=TRUE), 
              map_signif_level = TRUE,
              col="black")

```



Compare cell type proportions between birth (using FlowSorted.CordBloodCombined.450k and any probes) and age 1 (using FlowSorted.Blood.Epic with adultIDOL probes).
```{r boxplot_cbany_y1idol}
#substring sample id so that numbering is consistent between birth and age one
child_fscbc_ecc2_cb_df_nodups$sampleid <- substr(child_fscbc_ecc2_cb_df_nodups$sampleid, 1, 5)
child_fsbepic_ecc2_idol_y1_df$sampleid <- substr(child_fsbepic_ecc2_idol_y1_df$sampleid, 1, 5) 

#populate nRBC column in age1 group with NA
child_fsbepic_ecc2_idol_y1_df$nRBC <- NA

#rearrange columns so that they are in the same order
child_fsbepic_ecc2_idol_y1_df <- child_fsbepic_ecc2_idol_y1_df[ colnames(child_fscbc_ecc2_cb_df_nodups)]
#check column order
head(child_fsbepic_ecc2_idol_y1_df)
head(child_fscbc_ecc2_cb_df_nodups)

#subset age one data so that only samples with data at both birth and age one are plotted/tests
child_fsbepic_ecc2_idol_y1_df <- child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                           child_fscbc_ecc2_cb_df_nodups$sampleid, ] 
#check sizes
dim(child_fsbepic_ecc2_idol_y1_df)
dim(child_fscbc_ecc2_cb_df_nodups)

#check that samples are in same order
identical(child_fscbc_ecc2_cb_df_nodups$sampleid, 
           child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                           child_fscbc_ecc2_cb_df_nodups$sampleid,"sampleid"]) #true

#rowbind dataframes together
cby1 <- rbind(child_fscbc_ecc2_cb_df_nodups, child_fsbepic_ecc2_idol_y1_df)

#melt dataframe for plotting
cby1_long <- cby1 %>% melt(id = c("tissue", "sampleid", "method"))

#change class of variables for plotting
cby1_long$sampleid <- as.factor(cby1_long$sampleid)


#t-tests between birth and age one for cell types


#cd4t
cd4t <- t.test(child_fscbc_ecc2_cb_df_nodups[,"CD4T"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"CD4T"],
               alternative = "two.sided", 
               paired=TRUE)

#cd8t
cd8t <- t.test(child_fscbc_ecc2_cb_df_nodups[,"CD8T"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"CD8T"],
               alternative = "two.sided", 
               paired=TRUE)

#mono
mono <-  t.test(child_fscbc_ecc2_cb_df_nodups[,"Mono"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"Mono"],
               alternative = "two.sided", 
               paired=TRUE)

#NK cells
nk <-  t.test(child_fscbc_ecc2_cb_df_nodups[,"NK"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"NK"],
               alternative = "two.sided", 
               paired=TRUE)

#bcells
bcell <- t.test(child_fscbc_ecc2_cb_df_nodups[,"Bcell"], 
               child_fsbepic_ecc2_idol_y1_df[child_fsbepic_ecc2_idol_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"Bcell"],
               alternative = "two.sided", 
               paired=TRUE)


#graphing
ggplot(cby1_long, aes(x=as.factor(tissue),
                      y=value, 
                      group=sampleid, 
                      col=sampleid))  +
  geom_line() +
  geom_point(cby1_long %>% subset(variable=="nRBC"), 
             mapping=aes(x=as.factor(tissue), 
                         y=value, 
                         group=sampleid,
                         col=sampleid)) +
  facet_grid(. ~ variable) +
  theme(legend.position = "none") +
  theme_bw() +
  theme(legend.position ="none") +
  geom_signif(comparisons = list(c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P")),
              test="t.test", test.args=list(alternative="two.sided", paired=TRUE), 
              map_signif_level = TRUE,
              col="black")

```



Compare cell type proportions between birth (using FlowSorted.CordBloodCombined.450k and any probes) and age 1 (using FlowSorted.CordBloodCombined.450k any).
```{r boxplot_cbany_y1any}
#substring sample id so that numbering is consistent between birth and age one
child_fscbc_ecc2_cb_df_nodups$sampleid <- substr(child_fscbc_ecc2_cb_df_nodups$sampleid, 1, 5)
child_fscbc_ecc2_y1_df$sampleid <- substr(child_fscbc_ecc2_y1_df$sampleid, 1, 5) 

#populate nRBC column in age1 group with NA
child_fscbc_ecc2_y1_df$nRBC <- NA

#rearrange columns so that they are in the same order
child_fscbc_ecc2_y1_df <- child_fscbc_ecc2_y1_df[ colnames(child_fscbc_ecc2_cb_df_nodups)]
#check column order
head(child_fscbc_ecc2_y1_df)
head(child_fscbc_ecc2_cb_df_nodups)

#subset age one data so that only samples with data at both birth and age one are plotted/tests
child_fscbc_ecc2_y1_df <- child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                           child_fscbc_ecc2_cb_df_nodups$sampleid, ] 
#check sizes
dim(child_fscbc_ecc2_y1_df)
dim(child_fscbc_ecc2_cb_df_nodups)

#check that samples are in same order
identical(child_fscbc_ecc2_cb_df_nodups$sampleid, 
           child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                           child_fscbc_ecc2_cb_df_nodups$sampleid,"sampleid"]) #true

#rowbind dataframes together
cby1 <- rbind(child_fscbc_ecc2_cb_df_nodups, child_fscbc_ecc2_y1_df)

#melt dataframe for plotting
cby1_long <- cby1 %>% melt(id = c("tissue", "sampleid", "method"))

#change class of variables for plotting
cby1_long$sampleid <- as.factor(cby1_long$sampleid)


#t-tests between birth and age one for cell types


#cd4t
cd4t <- t.test(child_fscbc_ecc2_cb_df_nodups[,"CD4T"], 
               child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"CD4T"],
               alternative = "two.sided", 
               paired=TRUE)

#cd8t
cd8t <- t.test(child_fscbc_ecc2_cb_df_nodups[,"CD8T"], 
               child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"CD8T"],
               alternative = "two.sided", 
               paired=TRUE)

#mono
mono <-  t.test(child_fscbc_ecc2_cb_df_nodups[,"Mono"], 
               child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"Mono"],
               alternative = "two.sided", 
               paired=TRUE)

#NK cells
nk <-  t.test(child_fscbc_ecc2_cb_df_nodups[,"NK"], 
               child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"NK"],
               alternative = "two.sided", 
               paired=TRUE)

#bcells
bcell <- t.test(child_fscbc_ecc2_cb_df_nodups[,"Bcell"], 
               child_fscbc_ecc2_y1_df[child_fscbc_ecc2_y1_df$sampleid %in%
                                             child_fscbc_ecc2_cb_df_nodups$sampleid,"Bcell"],
               alternative = "two.sided", 
               paired=TRUE)


#graphing
ggplot(cby1_long, aes(x=as.factor(tissue),
                      y=value, 
                      group=sampleid, 
                      col=sampleid))  +
  geom_line() +
  geom_point(cby1_long %>% subset(variable=="nRBC"), 
             mapping=aes(x=as.factor(tissue), 
                         y=value, 
                         group=sampleid,
                         col=sampleid)) +
  facet_grid(. ~ variable) +
  theme(legend.position = "none") +
  theme_bw() +
  theme(legend.position ="none") +
  geom_signif(comparisons = list(c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P"), c("C", "P")),
              test="t.test", test.args=list(alternative="two.sided", paired=TRUE), 
              map_signif_level = TRUE,
              col="black")

```